---
title: "Causes of deaths 2011-2021, mod4"
date: "`r format(Sys.time(), 'Report generated on Date: %Y-%m-%d Time: %H:%M')`"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
---

      
```{r, include=FALSE}
source("R/000_setup.R")

wd_ofsp = "L:/UNISANTE_DESS/S_SUMAC/OFSP_2023/"
data_folder = paste0(wd_ofsp,"02_data/cause_of_death/")

cod_agg_pop_df = readRDS("savepoint/cod_agg_pop_df.RDS")
causes = c("Cardiovascular Diseases","External Causes","Infectious and Parasitic Diseases",
           "Mental and Neurological Disorders",
           "Neoplasms (Cancers)","No Specific Causes", "Respiratory Diseases", "Suicide")
age_classes = cod_agg_pop_df$age_class %>% unique()
res_list=load_results_mod4(age_classes, causes, save.date="20241114")

```

# Aim

We aim to
1. explore the mortality trends over time, by cause in Switzerland
2. assess excess mortality during 2020-2021, by cause
3. investigate the correlations between mortality causes (e.g., COVID-19 and other causes)

# Data

## Causes of deaths

## Reported deaths

* individual, with information about the calendar year, calendar week, exact age, canton and sex
* ICD-10 code of the cause of deaths, as well as secondary and other causes

```{r,  echo = FALSE}
cod_df0 = readRDS(paste0(code_root_path,"/data/cause_of_death2.rds"))
print(colnames(cod_df0))

```

### ICD-10 code


```{r,  echo = FALSE}
list = get_icd10()
icd10_chapter = list$icd10_chapter
icd10_chapter %>% flextable(cwidth = c(1,5))
```

### By cause and age, over calendar year

```{r,  echo = FALSE}


```

### Comparison with mortality reported by FSO



### Comparison with data from excess mortality



## Population

To predict the mortality risk by age group, we need to adjust for the population size.
We retrieve the population for each age group on the 1st of January of each year from FSO website and interpolate it to obtain the weekly population.

```{r,  fig.width=8,fig.height=8,echo = FALSE}
data.plot = cod_agg_pop_df %>% 
  dplyr::mutate(date=ISOweek2date(paste0(cal_year,"-W",ifelse(cal_week<10,paste0("0",cal_week),cal_week),"-1"))) %>% 
  dplyr::select(age_class,sex,cal_year,cal_week,date,n.pop) %>% unique() %>% 
  group_by(date,age_class,cal_week,cal_year,sex) %>% 
  dplyr::summarise(n=sum(n.pop),.groups="drop")
ggplot() +
  geom_point(data=data.plot %>% filter(cal_week==1),aes(x=date,y=n)) +
  geom_line(data=data.plot,aes(x=date,y=n)) +
  facet_grid(age_class~sex,scales="free") +
  #annotate("text",x=2020,y=0,label="") +
  theme_bw()
```


# Analysis

## Fitting mortality 2010-2019

For each cause and each age group, we use the weekly mortality data in 2010-2019 to predict the weekly mortality in 2020.
We assumed that the number of deaths $y_{i}(t)$ due to a given cause in given age group at time $t$ for sex $i$ followed a Poisson distribution:
$$y_{i}(t)\sim Poisson(\mu_{i}(t))$$

We model the changes of mortality risk over time using two Gaussian Processes (GPs).
We use an offset for the total population $N$ in the given age group:

$$\log\mu_{ijk} = \alpha + \beta_{F}\mathbb{1}_{i=F} + f(t) + g(t) + \log(N)$$
where $\alpha$ is the intercept,$\beta_F$ is the effect of being a female on the mortality and $N$ is the population.
The GP $f$ models the long-term trend in mortality over time while the GP $g$ represents the change in mortality over the weeks of the year.
We assume an exponential quadratic Kernel for $f$ and a periodic Kernel for $g$.

## Excess mortality

The model provides estimates of the weekly expected mortality by cause and age class in 2020.
We obtain the excess mortality by subtracting the expected mortality to the reported number of deaths.

# Results

## Fit 2010-2019

We only show fit for 70-79 and 80+, the two groups who gather most of the deaths.
The red dots represent the weekly number of deaths, the black lines the estimated mean number of deaths and the gray areas the 95% credibility intervals.

### 70-79

```{r,  fig.width=8,fig.height=8,echo = FALSE}

res_list$data_pred_week %>% 
  filter(variable=="deaths",age_class=="65-79") %>% 
  left_join(res_list$data_pred_week %>% 
              filter(variable=="obs_deaths",age_class=="65-79") %>% dplyr::select(cod_group,obs_deaths=est,cal_year,cal_week),
            by=c("cal_year","cal_week","cod_group")) %>% 
  ggplot() +
  geom_line(aes(x=date,y=est),col="black") +
  geom_ribbon(aes(x=date,ymin=lwb,ymax=upb),fill="black",alpha=0.15) +
  geom_point(aes(x=date,y=obs_deaths),col="red",alpha=0.5,size=0.8) +
  geom_vline(aes(xintercept=ymd("2020-01-01")))+
  facet_grid(cod_1word~age_class,scales="free") +
  theme_bw()
```

### 80+

```{r,  fig.width=8,fig.height=8,echo = FALSE}
res_list$data_pred_week %>% 
  filter(variable=="deaths",age_class=="80+") %>% 
  left_join(res_list$data_pred_week %>% 
              filter(variable=="obs_deaths",age_class=="80+") %>% dplyr::select(cod_group,obs_deaths=est,cal_year,cal_week),
            by=c("cal_year","cal_week","cod_group")) %>% 
  ggplot() +
  geom_line(aes(x=date,y=est),col="black") +
  geom_ribbon(aes(x=date,ymin=lwb,ymax=upb),fill="black",alpha=0.15) +
  geom_point(aes(x=date,y=obs_deaths),col="red",alpha=0.5,size=0.8) +
  geom_vline(aes(xintercept=ymd("2020-01-01")))+
  facet_grid(cod_1word~age_class,scales="free") +
  theme_bw()
```

## Expected mortality

### by year

```{r,  fig.width=8,fig.height=10,echo = FALSE}
res_list$data_pred_year %>% 
  filter(variable=="deaths",age_class %in% c("65-79","80+")) %>% 
  left_join(res_list$data_pred_year %>% 
              filter(variable=="obs_deaths",age_class %in% c("65-79","80+")) %>% dplyr::select(cod_group,age_class,obs_deaths=est,cal_year),
            by=c("cal_year","cod_group","age_class")) %>% 
  ggplot() +
  geom_line(aes(x=cal_year,y=est),col="black") +
  geom_ribbon(aes(x=cal_year,ymin=lwb,ymax=upb),fill="black",alpha=0.15) +
  geom_point(aes(x=cal_year,y=obs_deaths),col="red",alpha=0.5,size=2) +
  facet_grid(cod_1word~age_class,scales="free") +
  geom_vline(aes(xintercept=2019.5))+
  theme_bw()
```


### by phase

```{r,  fig.width=8,fig.height=10,echo = FALSE}
res_list$data_pred_phase %>% 
  filter(variable=="deaths",age_class %in% c("65-79","80+")) %>% 
  left_join(res_list$data_pred_phase %>% 
              filter(variable=="obs_deaths",age_class %in% c("65-79","80+")) %>% dplyr::select(cod_group,age_class,obs_deaths=est,covid_phase),
            by=c("covid_phase","cod_group","age_class")) %>% 
  filter(covid_phase>0) %>% 
  left_join(covid_phase %>% dplyr::select(covid_phase=phase,n_weeks,labels),by="covid_phase") %>% 
  dplyr::mutate(est=est/n_weeks,lwb=lwb/n_weeks,upb=upb/n_weeks,obs_deaths=obs_deaths/n_weeks) %>% 
  ggplot(aes(x=covid_phase,y=est,ymin=lwb,ymax=upb)) +
  geom_line(col="black") +
  geom_ribbon(fill="black",alpha=0.15) +
  geom_point(aes(y=obs_deaths),col="red",alpha=0.5,size=2) +
  facet_grid(cod_1word~age_class,scales="free") +
  scale_x_continuous(breaks=covid_phase$phase,labels=covid_phase$labels)+
  theme_bw()+
  ylab("Number of deaths by week")+
  theme(axis.text.x = element_text( angle = 45,
                                    hjust = 1,vjust=1,size = 10))
```

## Excess mortality by cause

### Absolute

```{r,  fig.width=8,fig.height=12,echo = FALSE}
res_list$data_pred_year %>% 
  filter(variable=="excess",cal_year>=2020) %>% 
  ggplot(aes(x=cod_1word,y=est,ymin=lwb,ymax=upb,fill=cal_year,group=factor(cal_year))) +
  geom_hline(yintercept=0,colour="grey50") +
  geom_col(position = position_dodge(width=0.5),
           width=0.5,alpha=.5) +
  # geom_point(position = position_dodge(width=0.5),
  #               colour="black",alpha=.8) +
  geom_errorbar(position = position_dodge(width=0.5),
                colour="black",width=0.5,alpha=.8) +
  facet_wrap(age_class~.,scales="free",ncol=2) +
  #scale_fill_discrete(guide="none") +
  labs(x="Cause",y="Absolute excess mortality") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45,hjust = 1))
```

### Relative

```{r,  fig.width=8,fig.height=12,echo = FALSE}
res_list$data_pred_year %>% 
  filter(variable=="rel_excess",cal_year>=2020) %>% 
  dplyr::mutate(lwb=ifelse(is.infinite(est),NA,lwb),
                upb=ifelse(is.infinite(est),NA,upb),
                est=ifelse(is.infinite(est),NA,est)) %>% 
  ggplot(aes(x=cod_1word,y=est,ymin=lwb,ymax=upb,fill=cal_year,group=factor(cal_year))) +
  geom_hline(yintercept=0,colour="grey50") +
  geom_col(position = position_dodge(width=0.5),
           width=0.5,alpha=.5) +
  # geom_point(position = position_dodge(width=0.5),
  #               colour="black",alpha=.8) +
  geom_errorbar(position = position_dodge(width=0.5),
                colour="black",width=0.5,alpha=.8) +
  facet_wrap(age_class~.,scales="free",ncol=2) +
  #scale_fill_discrete(guide="none") +
  labs(x="Cause",y="Absolute excess mortality") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45,hjust = 1))+
  scale_y_continuous(labels = scales::percent)
```

## Excess mortality by cause and phase


```{r,  fig.width=8,fig.height=8,echo = FALSE}
covid_phase %>% flextable(cwidth=c(1,1,2.5,1,1))
```


### Absolute excess mortality (without COVID)

The deaths due to COVID-19 are removed for a better readability.

```{r,  fig.width=8,fig.height=12,echo = FALSE}
res_list$data_pred_phase %>% 
  left_join(res_list$data_pred_phase %>% 
              filter(variable=="obs_deaths",age_class %in% c("65-79","80+")) %>% dplyr::select(cod_group,age_class,obs_deaths=est,covid_phase),
            by=c("covid_phase","cod_group","age_class")) %>% 
  filter(variable=="excess",covid_phase>0) %>% 
  ggplot(aes(x=cod_1word,y=est,ymin=lwb,ymax=upb,fill=covid_phase,group=factor(covid_phase))) +
  geom_hline(yintercept=0,colour="grey50") +
  geom_col(position = position_dodge(width=0.5),
           width=0.5,alpha=.5) +
  # geom_point(position = position_dodge(width=0.5),
  #               colour="black",alpha=.8) +
  geom_errorbar(position = position_dodge(width=0.5),
                colour="black",width=0.5,alpha=.8) +
  facet_grid(age_class~.,scales="free") +
  labs(x="Cause",y="Absolute excess mortality") +
  scale_fill_continuous(breaks=covid_phase$phase,labels=covid_phase$labels)+
  theme_bw() +
  theme(axis.text.x=element_text(angle=45,hjust = 1)) %>% 
  theme(axis.text.x = element_text( angle = 45,
                                    hjust = 1,vjust=1,size = 10))
```

### Relative excess mortality

Results are only shown for the categories where both the observed and the expected mortality is greater than 0.

```{r,  fig.width=8,fig.height=12,echo = FALSE}
res_list$data_pred_phase %>% 
  left_join(res_list$data_pred_phase %>% 
              filter(variable=="obs_deaths",age_class %in% c("65-79","80+")) %>% dplyr::select(cod_group,age_class,obs_deaths=est,covid_phase),
            by=c("covid_phase","cod_group","age_class")) %>% 
  filter(variable=="rel_excess",covid_phase>0) %>% 
  dplyr::mutate(lwb=ifelse(is.infinite(est)|is.nan(est),NA,lwb),
                upb=ifelse(is.infinite(est)|is.nan(est),NA,upb),
                est=ifelse(is.infinite(est)|is.nan(est),NA,est)) %>% 
  ggplot(aes(x=cod_1word,y=est,ymin=lwb,ymax=upb,fill=covid_phase,group=factor(covid_phase))) +
  geom_hline(yintercept=0,colour="grey50") +
  geom_col(position = position_dodge(width=0.5),
           width=0.5,alpha=.5) +
  # geom_point(position = position_dodge(width=0.5),
  #               colour="black",alpha=.8) +
  geom_errorbar(position = position_dodge(width=0.5),
                colour="black",width=0.5,alpha=.8) +
  facet_wrap(age_class~.,scales="free",ncol=2) +
  #scale_fill_discrete(guide="none") +
  scale_fill_continuous(breaks=covid_phase$phase,labels=covid_phase$labels)+
  labs(x="Cause",y="Absolute excess mortality") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45,hjust = 1))+
  scale_y_continuous(labels = scales::percent)
```

## Gaussian processes

### Long-term trend

```{r,  fig.width=8,fig.height=8,echo = FALSE}
res_list$year_GP %>% 
  group_by(cod_1word,age_class) %>%
  dplyr::mutate(ref=as.numeric(as.numeric(date)==min(as.numeric(date))),
                est_rel=exp(est-est[ref==1])-1,
                lwb_rel=exp(lwb-est[ref==1])-1,
                upb_rel =exp(upb - est[ref==1])-1) %>% ungroup() %>%
  ggplot(aes(x=date,y=est_rel)) +
  geom_ribbon(aes(ymin=lwb_rel,ymax=upb_rel),alpha=0.2) +
  geom_point(size=0.5)+
  geom_line()+
  facet_grid(cod_1word~age_class,scales="free") +
  scale_y_continuous(labels = scales::percent,name="Relative change in mortality risk") +
  theme_bw()+
  theme(axis.text.x=element_text(angle=45,hjust = 1))
```


### Periodic trend

```{r,  fig.width=8,fig.height=8,echo = FALSE}
res_list$week_GP %>% 
  group_by(cod_1word,age_class) %>%
  dplyr::mutate(ref=as.numeric(as.numeric(corr_date)==min(as.numeric(corr_date))),
                est_rel=exp(est-est[ref==1])-1,
                lwb_rel=exp(lwb-est[ref==1])-1,
                upb_rel =exp(upb - est[ref==1])-1) %>% ungroup() %>%
  ggplot(aes(x=corr_date,y=est_rel)) +
  geom_ribbon(aes(ymin=lwb_rel,ymax=upb_rel),alpha=0.2) +
  geom_point(size=0.5)+
  geom_line()+
  facet_grid(cod_1word~age_class,scales="free") +
  scale_y_continuous(labels = scales::percent,name="Relative change in mortality") +
  theme_bw()+
  theme(axis.text.x=element_text(angle=45,hjust = 1))
```

## Regression

### Sex differences

```{r,  fig.width=8,fig.height=6,echo = FALSE}
#mortality risk by sex
res_list$reg_effect %>%
  filter(var=="sex") %>%
  ggplot(aes(x=cod_1word,y=mean,ymin=`2.5%`,ymax=`97.5%`,col=age_class))+
  geom_pointrange(position=position_dodge(width=0.5))+
  geom_hline(aes(yintercept=0),lty=2)+
  scale_y_continuous(name="Mortality risk ratio (RR), women vs men",
                     limits=log(c(0.1,2)),
                     breaks=log(c(0.1,0.2,0.5,1,2,5,10)),#log(c(-0.9,-0.75,-0.5,0,1,2)+1),
                     labels = exp)+# labels = function(x){return(scales::percent(exp(x)-1))})+
  theme_bw()+
  theme(axis.text.x = element_text( angle = 45,
                                    hjust = 1,vjust=1,size = 10))
```
