---
title: "Causes of deaths 2011-2021"
date: "`r format(Sys.time(), 'Report generated on Date: %Y-%m-%d Time: %H:%M')`"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
---

      
```{r, include=FALSE}
source("R/000_setup.R")

wd_ofsp = "L:/UNISANTE_DESS/S_SUMAC/OFSP_2023/"
data_folder = paste0(wd_ofsp,"02_data/cause_of_death/")

cod_agg_pop_df = readRDS("savepoint/cod_agg_pop_df.RDS")
age_classes = cod_agg_pop_df$age_class %>% unique()
res_list=load_results_mod6(age_classes, save.date="20241218",mod="mod8")
causes = c("Cardiovascular Diseases","Respiratory Diseases", "Mental and Neurological Disorders",
             "Infectious and Parasitic Diseases",
             "Neoplasms (Cancers)","Suicide","External Causes",
             "Other Causes")
#for sex differences
res_list_mod4=load_results_mod4(age_classes, causes, save.date="20241212")

cod_agg_pop_df = cod_agg_pop_df %>%
  dplyr::mutate(date=ISOweek2date(paste0(cal_year,"-W",ifelse(cal_week<10,paste0("0",cal_week),cal_week),"-1"))) %>% 
  mutate(covid_phase = map2_dbl(date, list(covid_phase), function(d, phases) {
    phase <- phases %>%
      filter(d >= start_date & d <= end_date) %>%
      pull(phase)
    if (length(phase) == 0) NA_real_ else phase
  }))

```

# Aim

We aim to
1. explore the mortality trends over time, by cause in Switzerland
2. assess excess mortality during 2020-2021, by cause
3. investigate the correlations between mortality causes (e.g., COVID-19 and other causes)

# Data

## Causes of deaths

### 2020-2021

```{r,  fig.width=8,fig.height=8,echo = FALSE}
cod_ind_df=readRDS(paste0(code_root_path,"savepoint/cod_ind_df.RDS"))

causes2 = c("Cardiovascular Diseases","Infectious and Parasitic Diseases",
           "Respiratory Diseases", "Mental and Neurological Disorders",
           "COVID-19",
           "Neoplasms (Cancers)","Suicide","External Causes")

cod_ind_df %>% 
  filter(cal_year>=2020) %>% 
  group_by(outcome,cod_group) %>% 
  dplyr::summarise(n=n()) %>% ungroup() %>% 
  filter(!is.na(cod_group)) %>% 
  dplyr::mutate(outcome=factor(outcome,
                               levels=c("ENDG_U_CD_GES_T","GRUND_KRANK_GES_T","FOLGE_KRANK_GES_T","BEGLEIT_KRANK_A_GES_T","BEGLEIT_KRANK_B_GES_T"))) %>% 
  ggplot(aes(x=outcome,y=n,fill=cod_group)) +
  geom_bar(stat="identity")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Correlation: principal and primary

```{r,  fig.width=8,fig.height=8,echo = FALSE}
cod_ind_df %>%
  dplyr::mutate(outcome=factor(outcome,levels=c("ENDG_U_CD_GES_T","GRUND_KRANK_GES_T"),
                               labels=c("outcome1","outcome2"))) %>% 
  filter(!is.na(outcome)) %>% 
  pivot_wider(id_cols=ind_id,names_from="outcome",values_from="cod_group") %>% 
  group_by(outcome1,outcome2) %>% 
  dplyr::summarise(n=n(),.groups="drop_last") %>%
  dplyr::mutate(p=n/sum(n)) %>% ungroup() %>% 
  dplyr::mutate_at(c("outcome1","outcome2"),function(x) factor(x,levels=c(causes2,"Other Causes"))) %>% 
  ggplot(aes(x = outcome2, y = fct_rev(outcome1), fill = p)) +
  geom_tile() +                          # Creates the heatmap tiles
  geom_text(aes(label = scales::percent(p, accuracy = 1)),  # Add percentage labels
            color = "black", size = 2.5) +
  scale_fill_gradient(low = "lightyellow", high = "orangered1",
                      limits=c(0,0.2),
                      labels=scales::label_percent(accuracy = 1)) + # Adjust the color gradient
  labs(x = "Primary (GRUND_KRANK_GES_T)", y = "Principal (ENDG_U_CD_GES_T)", fill = "Count") + # Axis and legend labels
  theme_bw() +                      # Minimal theme for a clean look
  theme(axis.text.x = element_text(angle = 45, hjust = 0))+
  scale_x_discrete(position = "top")
```

### Correlation: principal and secondary

```{r,  fig.width=8,fig.height=8,echo = FALSE}
cod_ind_df %>%
  dplyr::mutate(outcome=factor(outcome,levels=c("ENDG_U_CD_GES_T","FOLGE_KRANK_GES_T"),
                               labels=c("outcome1","outcome2"))) %>% 
  filter(!is.na(outcome)) %>% 
  pivot_wider(id_cols=ind_id,names_from="outcome",values_from="cod_group") %>% 
  group_by(outcome1,outcome2) %>% 
  dplyr::summarise(n=n(),.groups="drop_last") %>%
  dplyr::mutate(p=n/sum(n)) %>% ungroup() %>% 
  dplyr::mutate_at(c("outcome1","outcome2"),function(x) factor(x,levels=c(causes2,"Other Causes"))) %>% 
  ggplot(aes(x = outcome2, y = fct_rev(outcome1), fill = p)) +
  geom_tile() +                          # Creates the heatmap tiles
  geom_text(aes(label = scales::percent(p, accuracy = 1)),  # Add percentage labels
            color = "black", size = 2.5) +
  scale_fill_gradient(low = "lightgreen", high = "lightblue",
                      labels=scales::label_percent(accuracy = 1)) + # Adjust the color gradient
  labs(x = "Secondary (FOLGE_KRANK_GES_T)", y = "Principal (ENDG_U_CD_GES_T)", fill = "Count") + # Axis and legend labels
  theme_bw() +                      # Minimal theme for a clean look
  theme(axis.text.x = element_text(angle = 45, hjust = 0))+
  scale_x_discrete(position = "top")
```

## Reported deaths

* individual, with information about the calendar year, calendar week, exact age, canton and sex
* ICD-10 code of the cause of deaths, as well as secondary and other causes

```{r,  echo = FALSE}
cod_df0 = readRDS(paste0(code_root_path,"/data/cause_of_death2.rds"))
print(colnames(cod_df0))
```

### ICD-10 code


```{r,  echo = FALSE}
list = get_icd10()
icd10_chapter = list$icd10_chapter
icd10_chapter %>% flextable(cwidth = c(1,5))
```

### By cause and age, over calendar year

```{r,  echo = FALSE}


```

### Comparison with mortality reported by FSO



### Comparison with data from excess mortality



## Population

To predict the mortality risk by age group, we need to adjust for the population size.
We retrieve the population for each age group on the 1st of January of each year from FSO website and interpolate it to obtain the weekly population.

```{r,  fig.width=8,fig.height=8,echo = FALSE}
data.plot = cod_agg_pop_df %>% 
  dplyr::mutate(date=ISOweek2date(paste0(cal_year,"-W",ifelse(cal_week<10,paste0("0",cal_week),cal_week),"-1"))) %>% 
  dplyr::select(age_class,sex,cal_year,cal_week,date,n.pop) %>% unique() %>% 
  group_by(date,age_class,cal_week,cal_year,sex) %>% 
  dplyr::summarise(n=sum(n.pop),.groups="drop")
ggplot() +
  geom_point(data=data.plot %>% filter(cal_week==1),aes(x=date,y=n)) +
  geom_line(data=data.plot,aes(x=date,y=n)) +
  facet_grid(age_class~sex,scales="free") +
  #annotate("text",x=2020,y=0,label="") +
  theme_bw()
```


# Analysis

## Fitting mortality 2010-2019

For each cause and each age group, we use the weekly mortality data in 2010-2019 to predict the weekly mortality in 2020.
We assumed that the number of deaths $y_{i}(t)$ due to a given cause in given age group at time $t$ for sex $i$ followed a Poisson distribution:
$$y_{i}(t)\sim Poisson(\mu_{i}(t))$$

We model the changes of mortality risk over time using two Gaussian Processes (GPs).
We use an offset for the total population $N$ in the given age group:

$$\log\mu_{ijk} = \alpha + \beta_{F}\mathbb{1}_{i=F} + f(t) + g(t) + \log(N)$$
where $\alpha$ is the intercept,$\beta_F$ is the effect of being a female on the mortality and $N$ is the population.
The GP $f$ models the long-term trend in mortality over time while the GP $g$ represents the change in mortality over the weeks of the year.
We assume an exponential quadratic Kernel for $f$ and a periodic Kernel for $g$.

## Excess mortality

The model provides estimates of the weekly expected mortality by cause and age class in 2020.
We obtain the excess mortality by subtracting the expected mortality to the reported number of deaths.

# Results

## Model fit

We only show fit for 65-79 and 80+, the two groups who gather most of the deaths.
The red dots represent the weekly number of deaths, the black lines the estimated mean number of deaths and the gray areas the 95% credibility intervals.

### By cause
#### 65-79

```{r,  fig.width=8,fig.height=12,echo = FALSE}
res_list$data_pred_week_cause %>% 
  filter(variable=="deaths") %>% 
  left_join(res_list$data_pred_week_cause %>% 
              filter(variable=="obs_deaths") %>% dplyr::select(obs_deaths=est,cal_year,cal_week,age_class,cod_group,pred),
            by=c("cal_year","cal_week","age_class","cod_group","pred")) %>% 
  filter(age_class=="65-79",pred=="poisson") %>% 
  ggplot() +
  geom_line(aes(x=date,y=est),col="black") +
  geom_ribbon(aes(x=date,ymin=lwb,ymax=upb),fill="black",alpha=0.15) +
  geom_point(aes(x=date,y=obs_deaths),col="red",alpha=0.5,size=0.8) +
  geom_vline(aes(xintercept=ymd("2020-01-01")))+
  facet_grid(cod_group~age_class,scales="free") +
  theme_bw()
```

#### 80+

```{r,  fig.width=8,fig.height=12,echo = FALSE}
res_list$data_pred_week_cause %>% 
  filter(variable=="deaths") %>% 
  left_join(res_list$data_pred_week_cause %>% 
              filter(variable=="obs_deaths") %>% dplyr::select(obs_deaths=est,cal_year,cal_week,age_class,cod_group,pred),
            by=c("cal_year","cal_week","age_class","cod_group","pred")) %>% 
  filter(age_class=="80+",pred=="poisson") %>% 
  ggplot() +
  geom_line(aes(x=date,y=est),col="black") +
  geom_ribbon(aes(x=date,ymin=lwb,ymax=upb),fill="black",alpha=0.15) +
  geom_point(aes(x=date,y=obs_deaths),col="red",alpha=0.5,size=0.8) +
  geom_vline(aes(xintercept=ymd("2020-01-01")))+
  facet_grid(cod_group~age_class,scales="free") +
  theme_bw()
```

### Aggregated over causes

Red: overall deaths. Blue: deaths without COVID-19.

```{r,  fig.width=8,fig.height=12,echo = FALSE}
res_list$data_pred_week %>% 
  filter(variable=="deaths") %>% 
  #add observed deaths
  left_join(res_list$data_pred_week %>% 
              filter(variable=="obs_deaths") %>% dplyr::select(obs_deaths=est,cal_year,cal_week,age_class,pred),
            by=c("cal_year","cal_week","age_class","pred")) %>% 
  #add covid deaths to observed deaths
  left_join(cod_agg_pop_df %>% filter(cod_group=="COVID-19") %>% 
              group_by(age_class,cal_year, cal_week) %>% 
              dplyr::summarise(n_covid=sum(n),.groups="drop"),by=c("age_class","cal_year","cal_week")) %>% 
  dplyr::mutate(obs_deaths_with_covid=obs_deaths+n_covid) %>% 
  filter(pred=="dispersed poisson") %>% 
  ggplot() +
  geom_line(aes(x=date,y=est),col="black") +
  geom_ribbon(aes(x=date,ymin=lwb,ymax=upb),fill="black",alpha=0.15) +
  geom_point(aes(x=date,y=obs_deaths),col="blue",alpha=0.5,size=0.8) +
  geom_point(aes(x=date,y=obs_deaths_with_covid),col="red",alpha=0.5,size=0.8) +
  geom_vline(aes(xintercept=ymd("2020-01-01")))+
  facet_grid(age_class~.,scales="free") +
  theme_bw()
```

## Expected mortality

### by year, by cause

```{r,  fig.width=8,fig.height=10,echo = FALSE}
res_list$data_pred_year_cause %>% 
  filter(variable=="deaths") %>% 
  #add observed deaths
  left_join(res_list$data_pred_year_cause %>% 
              filter(variable=="obs_deaths") %>% dplyr::select(obs_deaths=est,cal_year,age_class,cod_group,pred),
            by=c("cal_year","cod_group","age_class","pred")) %>% 
  filter(age_class %in% c("65-79","80+"),pred=="poisson") %>% 
  ggplot() +
  geom_line(aes(x=cal_year,y=est),col="black") +
  geom_ribbon(aes(x=cal_year,ymin=lwb,ymax=upb),fill="black",alpha=0.15) +
  geom_point(aes(x=cal_year,y=obs_deaths),col="red",alpha=0.5,size=2) +
  facet_grid(cod_group~age_class,scales="free") +
  geom_vline(aes(xintercept=2019.5))+
  theme_bw()
```

### by year, aggregated

```{r,  fig.width=8,fig.height=10,echo = FALSE}
res_list$data_pred_year %>% 
  filter(variable=="deaths") %>% 
  #add observed deaths
  left_join(res_list$data_pred_year %>% 
              filter(variable=="obs_deaths") %>% dplyr::select(obs_deaths=est,cal_year,age_class,pred),
            by=c("cal_year","age_class","pred")) %>% 
  #add covid deaths to observed deaths
  left_join(cod_agg_pop_df %>% filter(cod_group=="COVID-19") %>% 
              group_by(cal_year,age_class) %>% 
              dplyr::summarise(n_covid=sum(n),.groups="drop"),by=c("age_class","cal_year")) %>% 
  dplyr::mutate(obs_deaths_with_covid=obs_deaths+n_covid) %>% 
  filter(pred=="poisson") %>% 
  ggplot() +
  geom_line(aes(x=cal_year,y=est),col="black") +
  geom_ribbon(aes(x=cal_year,ymin=lwb,ymax=upb),fill="black",alpha=0.15) +
  geom_point(aes(x=cal_year,y=obs_deaths),col="blue",alpha=0.5,size=2) +
  geom_point(aes(x=cal_year,y=obs_deaths_with_covid),col="red",alpha=0.5,size=2) +
  facet_grid(age_class~.,scales="free") +
  geom_vline(aes(xintercept=2019.5))+
  theme_bw()
```


### by phase

```{r,  fig.width=8,fig.height=10,echo = FALSE}
res_list$data_pred_phase %>% 
  filter(variable=="deaths") %>% 
  #add observed deaths
  left_join(res_list$data_pred_phase %>% 
              filter(variable=="obs_deaths") %>% dplyr::select(obs_deaths=est,covid_phase,age_class,pred),
            by=c("covid_phase","age_class","pred")) %>% 
  #add covid deaths to observed deaths
  left_join(cod_agg_pop_df %>% filter(cod_group=="COVID-19") %>% 
              group_by(covid_phase,age_class) %>% 
              dplyr::summarise(n_covid=sum(n),.groups="drop"),by=c("age_class","covid_phase")) %>% 
  dplyr::mutate(obs_deaths_with_covid=obs_deaths+n_covid) %>% 
  filter(pred=="poisson",covid_phase>0) %>% 
  ggplot(aes(x=covid_phase,y=est,ymin=lwb,ymax=upb)) +
  geom_line(col="black") +
  geom_ribbon(fill="black",alpha=0.15) +
  geom_point(aes(y=obs_deaths),col="blue",alpha=0.5,size=2) +
  geom_point(aes(y=obs_deaths_with_covid),col="red",alpha=0.5,size=2) +
  facet_grid(age_class~.,scales="free") +
  scale_x_continuous(breaks=covid_phase$phase,labels=covid_phase$labels)+
  theme_bw()+
  ylab("Number of deaths by week")+
  theme(axis.text.x = element_text( angle = 45,
                                    hjust = 1,vjust=1,size = 10))
```

## Excess mortality

### Aggregated over causes, by age and year

```{r,  fig.width=8,fig.height=12,echo = FALSE}
res_list$data_pred_year %>% 
  filter(variable=="excess",pred=="poisson") %>% 
  ggplot(aes(x=cal_year,y=est,ymin=lwb,ymax=upb)) +
  geom_hline(yintercept=0,colour="grey50") +
  geom_col(position = position_dodge(width=0.5),
           width=0.5,alpha=.5) +
  # geom_point(position = position_dodge(width=0.5),
  #               colour="black",alpha=.8) +
  geom_errorbar(position = position_dodge(width=0.5),
                colour="black",width=0.5,alpha=.8) +
  facet_wrap(age_class~.,scales="free",ncol=2) +
  #scale_fill_discrete(guide="none") +
  labs(x="Cause",y="Absolute excess mortality") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45,hjust = 1))
```

### By causes and year, 80+

```{r,  fig.width=8,fig.height=12,echo = FALSE}
res_list$data_pred_year_cause %>% 
  left_join(cod_df,by=c("cod_group"="cod_full")) %>% 
  filter(variable=="excess",age_class=="80+",pred=="poisson") %>% 
  ggplot(aes(x=cal_year,y=est,ymin=lwb,ymax=upb)) +
  geom_hline(yintercept=0,colour="grey50") +
  geom_col(position = position_dodge(width=0.5),
           width=0.5,alpha=.5) +
  # geom_point(position = position_dodge(width=0.5),
  #               colour="black",alpha=.8) +
  geom_errorbar(position = position_dodge(width=0.5),
                colour="black",width=0.5,alpha=.8) +
  facet_wrap(cod_1word~.,scales="free",ncol=2) +
  #scale_fill_discrete(guide="none") +
  labs(x="Cause",y="Absolute excess mortality") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45,hjust = 1))
```

### By causes and age, 2020-21

```{r,  fig.width=8,fig.height=12,echo = FALSE}
res_list$data_pred_year_cause %>% 
  left_join(cod_df,by=c("cod_group"="cod_full")) %>% 
  filter(variable=="excess",cal_year>=2020,pred=="poisson") %>% 
  ggplot(aes(x=cod_1word,y=est,ymin=lwb,ymax=upb,fill=factor(cal_year),group=factor(cal_year))) +
  geom_hline(yintercept=0,colour="grey50") +
  geom_col(position = position_dodge(width=0.5),
           width=0.5,alpha=.5) +
  # geom_point(position = position_dodge(width=0.5),
  #               colour="black",alpha=.8) +
  geom_errorbar(position = position_dodge(width=0.5),
                colour="black",width=0.5,alpha=.8) +
  facet_wrap(age_class~.,scales="free",ncol=2) +
  scale_fill_manual(values=c("#009999", "#0000FF")) +
  labs(x="Cause",y="Absolute excess mortality") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45,hjust = 1))
```

### by phase

```{r,  fig.width=8,fig.height=15,echo = FALSE}
res_list$data_pred_phase_cause %>% 
  left_join(cod_df,by=c("cod_group"="cod_full")) %>% 
  filter(variable=="excess",covid_phase>0,pred=="poisson") %>% 
  ggplot(aes(x=cod_1word,y=est,ymin=lwb,ymax=upb,fill=factor(covid_phase),group=factor(covid_phase))) +
  geom_hline(yintercept=0,colour="grey50") +
  geom_col(position = position_dodge(width=0.8),
           width=0.8,alpha=.5) +
  # geom_point(position = position_dodge(width=0.5),
  #               colour="black",alpha=.8) +
  geom_errorbar(position = position_dodge(width=0.8),
                colour="black",width=0.8,alpha=.8) +
  facet_wrap(age_class~.,scales="free",ncol=1) +
  scale_fill_manual(values=viridis_pal()(9)) +
  labs(x="Cause",y="Absolute excess mortality") +
  theme_bw() +
  theme(axis.text.x=element_text(angle=45,hjust = 1),
        legend.position = "bottom")
```

## Gaussian processes

### Long-term trend

```{r,  fig.width=8,fig.height=8,echo = FALSE}
res_list$year_GP %>% 
  left_join(cod_df,by=c("cod_group"="cod_full")) %>% 
  group_by(cod_1word,age_class) %>%
  dplyr::mutate(ref=as.numeric(as.numeric(date)==min(as.numeric(date))),
                est_rel=exp(est-est[ref==1])-1,
                lwb_rel=exp(lwb-est[ref==1])-1,
                upb_rel =exp(upb - est[ref==1])-1) %>% ungroup() %>%
  ggplot(aes(x=date,y=est_rel)) +
  geom_vline(xintercept=as.Date("2020-01-01"),lty=2)+
  geom_ribbon(aes(ymin=lwb_rel,ymax=upb_rel),alpha=0.2) +
  geom_point(size=0.5)+
  geom_line()+
  facet_grid(cod_1word~age_class,scales="fixed") +
  scale_y_continuous(labels = scales::percent,name="Relative change in mortality risk") +
  theme_bw()+
  theme(axis.text.x=element_text(angle=45,hjust = 1))
```

### Periodic trend

```{r,  fig.width=8,fig.height=8,echo = FALSE}
res_list$week_GP %>% 
  left_join(cod_df,by=c("cod_group"="cod_full")) %>% 
  group_by(cod_1word,age_class) %>%
  dplyr::mutate(ref=as.numeric(as.numeric(corr_date)==min(as.numeric(corr_date))),
                est_rel=exp(est-est[ref==1])-1,
                lwb_rel=exp(lwb-est[ref==1])-1,
                upb_rel =exp(upb - est[ref==1])-1) %>% ungroup() %>%
  ggplot(aes(x=as.Date(corr_date),y=est_rel)) +
  geom_ribbon(aes(ymin=lwb_rel,ymax=upb_rel),alpha=0.2) +
  geom_point(size=0.5)+
  geom_line()+
  facet_grid(cod_1word~age_class,scales="fixed") +
  scale_x_date(date_labels ="%b",name="")+
  scale_y_continuous(labels = scales::percent,name="Relative change in mortality") +
  theme_bw()+
  theme(axis.text.x=element_text(angle=45,hjust = 1))
```


## Correlation

```{r,  fig.width=8,fig.height=6,echo = FALSE}
#mortality risk by sex
cod_order = res_list$Sigma_mat %>% dplyr::select(cod_group_id,cod_group) %>% unique() %>% 
  left_join(cod_df %>% dplyr::select(cod_group=cod_full,cod_1word=cod_1word),by=c("cod_group")) %>% 
  arrange(cod_group_id) %>% pull(cod_1word)

res_list$Sigma_mat %>% 
  left_join(cod_df %>% dplyr::select(cod_group=cod_full,cod_1word=cod_1word),by=c("cod_group")) %>% 
  left_join(cod_df %>% dplyr::select(cod_group2=cod_full,cod_1word2=cod_1word),by=c("cod_group2")) %>% 
  dplyr::mutate(cod_1word = factor(cod_1word, levels=cod_order),
                cod_1word2 = factor(cod_1word2, levels=cod_order)) %>% 
  filter(cod_group_id<cod_group_id2) %>% 
  ggplot(aes(x=cod_1word,y=est,ymin=lwb,ymax=upb,col=age_class))+
  geom_hline(yintercept = 0,lty=1)+
  geom_pointrange(position=position_dodge(width=0.5))+
  scale_y_continuous(limits = c(-1,1))+
  facet_grid(cod_1word2~.)+
  theme_bw()+
  theme(axis.text.x=element_text(angle=45,hjust = 1))
```

```{r,  fig.width=8,fig.height=6,echo = FALSE}
#mortality risk by sex
cod_order = res_list$Sigma_mat %>% dplyr::select(cod_group_id,cod_group) %>% unique() %>% 
  left_join(cod_df %>% dplyr::select(cod_group=cod_full,cod_1word=cod_1word),by=c("cod_group")) %>% 
  arrange(cod_group_id) %>% pull(cod_1word)
res_list$Sigma_mat %>% 
  left_join(cod_df %>% dplyr::select(cod_group=cod_full,cod_1word=cod_1word),by=c("cod_group")) %>% 
  left_join(cod_df %>% dplyr::select(cod_group2=cod_full,cod_1word2=cod_1word),by=c("cod_group2")) %>% 
  dplyr::mutate(cod_1word = factor(cod_1word, levels=cod_order),
                cod_1word2 = factor(cod_1word2, levels=cod_order)) %>% 
  rowwise() %>% 
  dplyr::mutate(est_cri = paste0(scales::percent(est, accuracy = 1),"\n",
                                 "[",scales::percent(lwb, accuracy = 1),",",
                                 scales::percent(upb, accuracy = 1),"]")) %>% 
  filter(cod_group_id<cod_group_id2,age_class=="80+") %>% 
  ggplot(aes(x = cod_1word, y = fct_rev(cod_1word2), fill = abs(est))) +
  geom_tile() +
  geom_text(aes(label = est_cri),
            color = "black", size = 3) +
  scale_fill_gradient(low = "lightyellow", high = "orangered1",limits=c(0,1),
                      name="Correlation",
                      labels=scales::label_percent(accuracy = 1)) +
  labs(x = "", y = "", fill = "Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0))+
  scale_x_discrete(position = "top")
```

## Regression

### Sex differences

```{r,  fig.width=8,fig.height=6,echo = FALSE}
#mortality risk by sex
# res_list_mod4$reg_effect %>%
#   filter(var=="sex") %>%
res_list$sex_effect %>%
  filter(sex=="F") %>%
  left_join(cod_df,c("cod_group"="cod_full")) %>% 
  ggplot(aes(x=cod_1word,y=mean,ymin=`2.5%`,ymax=`97.5%`,col=age_class))+
  geom_pointrange(position=position_dodge(width=0.5))+
  geom_hline(aes(yintercept=0),lty=2)+
  scale_y_continuous(name="Mortality risk ratio (RR), women vs men",
                     limits=log(c(0.1,2.1)),
                     breaks=log(c(0.1,0.2,0.5,1,2,5,10)),#log(c(-0.9,-0.75,-0.5,0,1,2)+1),
                     labels = exp)+
  theme_bw()+
  theme(axis.text.x = element_text( angle = 45,
                                    hjust = 1,vjust=1,size = 10))
```

### Regional differences

```{r,  fig.width=8,fig.height=6,echo = FALSE}

res_list$nuts_effect %>%
  left_join(cod_df,c("cod_group"="cod_full")) %>%
  ggplot(aes(x=cod_1word,y=mean,ymin=`2.5%`,ymax=`97.5%`,col=age_class))+
  geom_pointrange(position=position_dodge(width=0.5))+
  geom_hline(aes(yintercept=0),lty=2)+
  scale_y_continuous(name="Mortality risk ratio (RR), women vs men",
                     limits=log(c(0.1,2.1)),
                     breaks=log(c(0.1,0.2,0.5,1,2,5,10)),#log(c(-0.9,-0.75,-0.5,0,1,2)+1),
                     labels = exp)+
    facet_wrap(.~NUTS2_name)+
  theme(axis.text.x = element_text( angle = 45,
                                    hjust = 1,vjust=1,size = 10))
```

```{r,  fig.width=8,fig.height=6,echo = FALSE}

res_list$nuts_effect %>%
  left_join(cod_df,c("cod_group"="cod_full")) %>%
  ggplot(aes(x=cod_1word,y=mean,ymin=`2.5%`,ymax=`97.5%`,col=NUTS2_name))+
  geom_pointrange(position=position_dodge(width=0.5))+
  geom_hline(aes(yintercept=0),lty=2)+
  scale_y_continuous(name="Mortality risk ratio (RR)",
                     limits=log(c(0.1,2.1)),
                     breaks=log(c(0.1,0.2,0.5,1,2,5,10)),#log(c(-0.9,-0.75,-0.5,0,1,2)+1),
                     labels = exp)+
    facet_wrap(.~age_class)+
  theme(axis.text.x = element_text( angle = 45,
                                    hjust = 1,vjust=1,size = 10))
```



## Correlation with COVID-19

### Plot

For 80+.

```{r,  fig.width=8,fig.height=10,echo = FALSE}
#aggregate
df = cod_agg_pop_df %>%
  filter(age_class=="80+") %>% 
  dplyr::mutate(date=ISOweek2date(paste0(cal_year,"-W",ifelse(cal_week<10,paste0("0",cal_week),cal_week),"-1"))) %>% 
  group_by(age_class,cal_year,cal_week,cod_group,date) %>%
  dplyr::summarise(n=sum(n),
                   n.pop=sum(n.pop)) %>% ungroup() %>%
  filter(cal_year>=2020) %>% dplyr::select(date,cod_group,n)  %>% 
  dplyr::filter(cod_group %in% c("Cardiovascular Diseases","Respiratory Diseases",
                                 "Mental and Neurological Disorders","COVID-19")) %>% #"Neoplasms (Cancers)"
dplyr::mutate(cod_group = factor(cod_group,levels=causes2_df$cod_group,
                                   labels=causes2_df$cod_group_label))

df %>% 
  ggplot(aes(x=date,y=n))+
  geom_point()+
  facet_wrap()

# Reshape the data so that each cod_group becomes a column
reshaped_df <- pivot_wider(df, names_from = cod_group, values_from = n)
# Gather data for scatter plot pairs
pairwise_data <- combn(colnames(reshaped_df)[-1], 2, simplify = FALSE, FUN = function(pair) {
  X <- reshaped_df[[pair[1]]]
  Y <- reshaped_df[[pair[2]]]
  date <- reshaped_df[["date"]]
  cor_value <- cor(X, Y, use = "complete.obs")
  data.frame(
    X = X,
    Y = Y,
    date = date,
    Pair = paste(pair[1], "vs", pair[2], sep = " "),
    Correlation = paste("Corr =", round(cor_value, 2))
  )
}) %>% bind_rows()

# Plot scatter plots with facet_wrap
ggplot(pairwise_data, aes(x = X, y = Y)) +
  geom_point(color = "blue", size = 3) +          # Scatter points
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Regression line
  facet_wrap(~ Pair, scales = "free",ncol=2) +          # Facet by pairs
  geom_text(aes(label = Correlation), x = Inf, y = -Inf, 
            hjust = 1.1, vjust = -1.1, inherit.aes = FALSE) +  # Add correlation text
  theme_bw() +
  labs(title = "Scatter Plots with Regression Line and Correlation",
       x = "Deaths of Cause 1",
       y = "Deaths of Cause 2")
```

```{r,  fig.width=8,fig.height=10,echo = FALSE}
pairwise_data %>% 
  pivot_longer(cols=c("X",Y),names_to = "variable",values_to = "n") %>% 
  ggplot(aes(x =date,y=n,col=variable)) +
  geom_line()+
  facet_wrap(~ Pair, scales = "free",ncol=2) +          # Facet by pairs
  geom_text(aes(label = Correlation), x = Inf, y = -Inf, 
            hjust = 1.1, vjust = -1.1, inherit.aes = FALSE) +  # Add correlation text
  theme_bw() +
  labs(title = "Scatter Plots with Regression Line and Correlation",
       x = "Value of Cod Group 1",
       y = "Value of Cod Group 2")
```

### with lags

```{r,  fig.width=8,fig.height=10,echo = FALSE}
calculate_lagged_correlation <- function(x, y, lags) {
  correlations <- map_dfr(lags, function(lag) {
    if (lag < 0) {
      lagged_x <- x[(1 - lag):length(x)]
      lagged_y <- y[1:(length(y) + lag)]
    } else if (lag > 0) {
      lagged_x <- x[1:(length(x) - lag)]
      lagged_y <- y[(1 + lag):length(y)]
    } else {
      lagged_x <- x
      lagged_y <- y
    }
    cor_value <- cor(lagged_x, lagged_y, use = "complete.obs")
    n <- length(lagged_x)  # Sample size for this lag
    fisher_z <- atanh(cor_value)  # Fisher's z-transformation
    se <- 1 / sqrt(n - 3)  # Standard error of Fisher's z
    z_low <- fisher_z - 1.96 * se  # Lower bound of Fisher's z
    z_high <- fisher_z + 1.96 * se  # Upper bound of Fisher's z
    ci_low <- tanh(z_low)  # Transform back to correlation
    ci_high <- tanh(z_high)  # Transform back to correlation
    data.frame(Lag = lag, Correlation = cor_value, CI_Low = ci_low, CI_High = ci_high)
  })
  return(correlations)
}

# Compute correlations for all pairs of cod_group values with lags
lags <- -10:10
correlations_by_pair <- combn(colnames(reshaped_df)[-1], 2, simplify = FALSE, FUN = function(pair) {
  x <- reshaped_df[[pair[1]]]
  y <- reshaped_df[[pair[2]]]
  correlations <- calculate_lagged_correlation(x, y, lags)
  correlations$Pair <- paste(pair[1], "vs", pair[2], sep = " ")
  return(correlations)
}) %>% bind_rows()

# Plot lagged correlations with uncertainty intervals
ggplot(correlations_by_pair, aes(x = Lag, y = Correlation, group = Pair)) +
  geom_hline(yintercept=0)+
  geom_ribbon(aes(ymin = CI_Low, ymax = CI_High), fill = "black", alpha = 0.1) +  # Confidence interval ribbon
  geom_line(color = "blue", size = 1) +           # Correlation line
  geom_point(size = 3, color = "red") +           # Points for correlation values
  facet_wrap(~ Pair, scales = "fixed",ncol=2) +
  geom_vline(xintercept = 0,lty=2)+
  theme_bw() +
  ylim(c(-1,1)) +
  labs(title = "Lagged Correlations Between Cod Groups with Confidence Intervals",
       x = "Lag (week)",
       y = "Correlation")
```

