---
title: "COVID-19 disrupted patterns of cause-specific mortality in Switzerland"
date: "`r format(Sys.time(), 'Report generated on Date: %Y-%m-%d Time: %H:%M')`"
author: "Anthony Hauser* (1), Moritz Wagner (2), Anne Laube (2), Julien Riou (1)"
output:
    word_document: 
      reference_docx: template.docx
      number_sections: true
      df_print: paged
bibliography: bibliography.bib
csl: ieee.csl
---

(1) Department of Epidemiology and Health Systems, Unisanté, Center for Primary Care and Public Health and University of Lausanne, Lausanne, Switzerland;<w:br/>
(2) Federal Office of Public Health, Bern, Switzerland;<w:br/>
(*) Corresponding author: anthony.hauser@unisante.ch

```{r setup, include=TRUE}
#Setup
Sys.setlocale("LC_TIME", "C")
wd = getwd()
code_root_path = paste0(strsplit(wd, split="/manuscript")[[1]][1],"/")
print(code_root_path)
source(paste0(code_root_path,"R/000_setup.R"))
knitr::opts_chunk$set(echo = FALSE, warning =FALSE, message=FALSE)

wd_ofsp = "L:/UNISANTE_DESS/S_SUMAC/OFSP_2023/"
data_folder = paste0(wd_ofsp,"02_data/cause_of_death/")
save.date="20241218"
```


```{r, include= TRUE}
#Individual dataset
cod_ind_df = readRDS(paste0(code_root_path,"savepoint/cod_ind_df.RDS"))

causes2 = c("Cardiovascular Diseases","Infectious and Parasitic Diseases",
           "Respiratory Diseases", "Mental and Neurological Disorders",
           "COVID-19",
           "Neoplasms (Cancers)","Suicide","External Causes")

causes2_df = data.frame(cod_group=c("Cardiovascular Diseases","Infectious and Parasitic Diseases",
                                    "Respiratory Diseases", "Mental and Neurological Disorders",
                                    "COVID-19",
                                    "Neoplasms (Cancers)","Suicide","External Causes","Other Causes"),
                        cod_group_label=c("Cardiovascular","Infectious/Parasitic",
                                    "Respiratory", "Mental/Neurological",
                                    "COVID-19",
                                    "Cancers","Suicide","External","Other"),
                        order=c(1,7,4,3,9,2,6,5,8),
                        example = c("Ischaemic heart disease, heart attack", "Sepsis",
                                   "Pulmonary disease, pneumonia, influenza", "Dementia, Alzheimer, Parkinson",
                                   "COVID-19",
                                   "Lung, breast, prostate neoplasms","Intentional self-harm/poisoning","Fall, vehicule accident",
                                   "Unspecified, diabetes, senility, organ disease")) %>%
  #add chapter of cod-10
  left_join(cod_ind_df %>%
              filter(outcome=="ENDG_U_CD_GES_T") %>%
              dplyr::select(cod_group,chapter) %>% unique() %>%
              arrange(cod_group,chapter) %>%
              group_by(cod_group) %>%
              dplyr::summarise(chapter=paste(chapter, collapse=", ")),by="cod_group") %>%
  arrange(order)
```

```{r, include=FALSE}
#Aggregated dataset and results
add_apo = function(x){format(x, big.mark=",")}

cod_agg_pop_df = readRDS(paste0(code_root_path,"savepoint/cod_agg_pop_df.RDS"))
age_classes = cod_agg_pop_df$age_class %>% unique()
setwd(code_root_path)
res_list = load_results_mod6(age_classes, save.date=save.date,mod="mod8")

cod_agg_pop_df = cod_agg_pop_df %>%
  dplyr::mutate(date=ISOweek2date(paste0(cal_year,"-W",ifelse(cal_week<10,paste0("0",cal_week),cal_week),"-1"))) %>% 
  mutate(covid_phase = map2_dbl(date, list(covid_phase), function(d, phases) {
    phase <- phases %>%
      filter(d >= start_date & d <= end_date) %>%
      pull(phase)
    if (length(phase) == 0) NA_real_ else phase
  }))
```


# Abstract

```{r, include=FALSE}

```

**Introduction**
  
  
**Methods and Findings**
  
  
**Conclusion**
  
  
\newpage

# Introduction

Link between covid and other causes
direct: 
indirect: measures -> suicides, respiratory diseases, accident

-> correlation of mortality causes
  -> important to analyse them simultaneously (as increase of a cause can lead to decrease of another cause, and thus no all-cause excess)
  
Disruption of mortality patterns offers a unique opportunity to explore the nature/origin of correlation between causes (which almost all usually peaks during winter but why? is it due to the conditions? to resp virus? other?)

Gap: cause-specific mortality during covid


Objectives

\newpage

# Methods

## Data

We received individual-level data on cause-specific deaths for the years 2011–2021 from the Federal Statistical Office (FSO), which we aggregated by age group (0–17, 18–39, 40–64, 65–79, 80+), sex, NUTS-2 region, calendar week, and calendar year.
The data include information recorded in the death certificate—such as the immediate, underlying, and contributing causes of death—as well as the official underlying cause of death, which is determined by the FSO based on the coding rules of the International Classification of Diseases, 10th Revision (ICD-10).
All causes of death are coded using ICD-10 and were grouped into nine categories for analysis: Cardiovascular diseases (Chapter IX), Cancers (Chapter II), Mental and neurological disorders (Chapters V and VI), Respiratory diseases (Chapter X), External causes (Chapter XX), Suicide (Chapter XX), Infectious and parasitic diseases (Chapter I), Other causes (Chapters III, IV, VII, VIII, XI–XIX, XXI, XXII), and COVID-19 (Chapter XXII).
Population data by sex, NUTS-2 region, and age group for the years 2011–2021 were also provided by the Federal Statistical Office (FSO). We applied linear interpolation to obtain weekly population estimates.

<!-- In Switzerland, the internal cause-specific mortality database records detailed information on the circumstances of each death and the characteristics of the deceased. -->
<!-- For every death of a Swiss resident, a physician completes a death certificate specifying the immediate, underlying, and contributing causes of death. -->
<!-- The death certificate is then transmitted to the Federal Statistical Office (FSO), where information is anonymized, coded, and entered into the national mortality database -->
<!-- The official underlying cause of death is determined according to the coding rules of the International Classification of Diseases, 10th Revision (ICD-10). -->
<!-- Additional variables — such as age, canton of residence, calendar week, and year of death — are also recorded. -->


## Model

<!-- We modeled the weekly number of deaths by cause for each age group using a Poisson distribution with a multivariate structure.   -->
<!-- The model accounts for changes in population size $N$ over time via an offset term, and includes fixed effects $\beta$ and $\gamma$ for differences in mortality risk by sex and region, respectively. -->
<!-- Temporal variation is captured by two components: a seasonal trend modeled with a periodic Gaussian Process (GP) $f(t)$, and a long-term trend modeled using a GP with an exponentiated quadratic (squared exponential) kernel $g(t)$.   -->
<!-- To account for the correlation between mortality risks from different causes, the model incorporates a multivariate random effect $\epsilon$. -->
<!-- The model reads as follow: -->
<!-- $$y_{i}(t)\sim Poisson(\mu_{i}(t))$$ -->
<!-- with, -->
<!-- $$\log\mu_{irs}(t) = \alpha_i + \beta_{ir} + \gamma_{is} + f(t) + g(t) + \epsilon_i + \log N_{irst}$$ -->
<!-- where $y_{irs}(t) $ is the observed number of deaths due to cause $i$ at week $t$ in region $r$ and sex $s$. -->
<!-- The parameters $\alpha_i$ represents the baseline relative mortality risk. -->

We modeled the weekly number of deaths by cause for each age group using a Poisson distribution with a multivariate structure.  
The model accounts for changes in population size $ N $ over time via an offset term, and includes fixed effects $ \beta $ and $ \gamma $ for differences in mortality risk by region and sex, respectively.  
Temporal variation is captured by two components: a seasonal trend modeled with a periodic Gaussian Process (GP) $ f(t) $, and a long-term trend modeled using a GP with an exponentiated quadratic (squared exponential) kernel $ g(t) $.  
To account for the correlation between mortality risks across causes, the model incorporates a multivariate random effect $ \epsilon \sim \mathcal{N}(0,\Sigma)$, where $\Sigma$ is the correlation matrix 

The model is specified as follows:
$$
y_i(t) \sim \text{Poisson}(\mu_i(t))
$$
with
$$
\log \mu_{irs}(t) = \alpha_i + \beta_{ir} + \gamma_{is} + f(t) + g(t) + \epsilon_i + \log N_{irst},
$$
where $ y_{irs}(t)$ denotes the observed number of deaths due to cause \( i \) in week \( t \), region \( r \), and sex \( s \), and \( N_{irst} \) is the corresponding population at risk.
<!-- where $ y_{irs}(t) $ denotes the observed number of deaths due to cause $ i $ at week $ t $, in region $ r $ and sex $ s $; $ \alpha_i $ is the baseline log mortality risk for cause $ i $; $ \beta_{ir} $ and $ \gamma_{is} $ are age-specific fixed effects for region and sex, respectively; $ f(t) $ captures the seasonal variation; $ g(t) $ represents the long-term trend; $ \epsilon \sim \mathcal{N}(0,\Sigma)$ is a multivariate random effect capturing correlation across causes (with correlation matrix $\Sigma$); and $ N_{irst} $ is the population at risk, included as a log offset. -->


The model was run independently for each age group and fitted to the weekly number of cause-specific deaths $y_{irs}(t)$ observed between 2011 and 2019.
It was then used to estimate the weekly \textit{expected number of deaths}  $\tilde{y}_{irs}(t)$ through the end of 2021, defined as draws from the \textit{posterior predictive distribution} after removing the random effect $\epsilon$ from the log mortality risk $\mu_{irs}(t)$.  
We excluded $\epsilon$ because it is designed to capture \textit{excess mortality}—that is, unexplained or anomalous variation beyond structured trends.  
Excess mortality $E_{irs}(t)$ was then defined as the difference between observed and expected deaths:
$$E_{irs}(t) := y_{irs}(t) - \tilde{y}_{irs}(t).$$
where $y_{irs}(t)$ is the observed number of deaths and $\tilde{y}_{irs}(t)$ the predicted number under baseline conditions.  
We also report aggregated estimates of excess mortality, such as \textit{all-cause excess mortality} (aggregated over causes of death) or aggregated over time period.

## Further analysis

We examined how excess deaths from specific causes were related over time to excess deaths from COVID-19 during 2020–21.
To do this, we computed partial correlations between posterior samples of cause-specific excess deaths and COVID-19 deaths, while controlling for excess deaths from other respiratory diseases.
This adjustment aimed to separate the association with COVID-19 from that with other circulating respiratory pathogens.
We also explored the effect of time lags, by shifting the time series of cause-specific excess deaths up to 8 weeks forward and backward, to assess whether the strength of association varied depending on timing.
We performed the same analysis to assess the relationship between cause-specific excess deaths and excess deaths due to (non-COVID) respiratory diseases, this time adjusting for COVID-19 mortality.


Q: where to describe covid-19 phase in switzerland? Introduction?
\newpage

# Results

## Distribution of deaths

In Switzerland, deaths are heavily concentrated among older adults: 88% occurred in individuals aged 65 and over, while fewer than 2% were among those under 40 in 2020-21 (Fig 1A).
Cardiovascular diseases were the leading cause of death overall, followed by cancers and mental or neurological disorders, though the distribution varied by age group  (Fig 1B).
Among individuals aged 80 and above, cardiovascular disease accounted for XX% of all deaths.
In contrast, for those aged 40–79, cancer was the predominant cause, responsible for approximately 40% of deaths.
The relative impact of COVID-19 was greatest in older age groups, contributing to XX% of deaths among those aged 65–79 and XX% among those aged 80 and over.

## Prepandemic mortality trends

The model successfully captures both the seasonal and long-term trends in cause-specific mortality observed between 2011 and 2019 (Figure 2A for individuals aged 80+, and Figures SXX–SXX for other age groups).
We observe notable similarities in mortality trends across different causes, both in the expected seasonal patterns and in intermittent episodes of excess mortality that often occur synchronously (e.g., during the winters of 2014/15 and 2016/17).
For the 80+ age group, Figure 2B highlights this synchrony in timing, showing that the expected mean mortality peak (in black) occurs consistently across causes between January and February.
During the pre-pandemic period (2011–2019), observed peak timings (colored circles) align closely with these expectations.
However, during the pandemic years, we observe a marked shift: mortality peaks occur earlier than expected (colored squares).
In addition, the model reveals strong positive correlations in cause-specific excess mortality, mostly among individuals aged 80 and over (Fig 2C).
Specifically, excess deaths from respiratory diseases, mental or neurological disorders, and cardiovascular diseases are highly correlated, with pairwise correlations exceeding 80%.

## COVID-19 mortality

The model also captures all-cause mortality, obtained by aggregating deaths across causes.
Among individuals aged 65 and over, the impact of the first two COVID-19 waves—-spring 2020 and fall/winter 2020–21—-clearly exceeded any episodes of excess mortality observed between 2011 and 2019 (Figure 3A).
Cumulative excess mortality highlights the substantial increase in deaths during these waves, followed by a gradual mortality deficit in the subsequent weeks (Figure 3B).
When excluding deaths attributed to COVID-19, we also observe that mortality from other causes was lower than expected in 2020–21 for people aged 65 and over.

## Cause-specific excess mortality over time

Among individuals aged below 64, almost all cause-specific mortality remained within expected levels, suggesting no evidence of excess mortality during 2020–21.
A slight excess in suicides was observed among 0-17 (X [XX–XX], +XX%), but the small number of deaths (43) limits interpretation.
In contrast, among individuals aged 65–79, a deficit in deaths due to cancers and mental or neurological disorders was observed in 2020–21, with approximately 5% fewer deaths than expected.
The largest decrease occurred during the Alpha wave in January–February 2021.
Mortality due to cardiovascular diseases in this group remained within the expected range, although a localized excess was noted during the autumn 2020 wave (+XX%).

Similar patterns were seen in the 80+ age group.
A marked decrease in mortality from mental and neurological disorders was observed in 2020–21 (−XX%), with levels below expectation in all phases except the autumn 2020 wave, when a slight increase occurred.
The most pronounced decrease was again during the Alpha wave (−XX%).
For cancer, the mortality deficit was concentrated almost entirely during the Alpha wave.
Cumulative mortality due to cardiovascular diseases among the 80+ was within expected levels overall, but showed substantial variation across phases. Excess mortality was concentrated during COVID-19 waves, while mortality deficits were more common at the beginning of both 2020 and 2021.
Finally, we observed a consistent deficit in deaths due to respiratory diseases (excluding COVID-19) across all age groups, with the magnitude of the deficit increasing with age.

To better understand cancer-related mortality patterns, we examined deaths among individuals with a cancer diagnosis, stratified by cancer types grouped according to baseline survival.
Unlike previous analyses based on the underlying cause of death, we also included individuals with cancer who died from other causes. The number of people dying with a cancer diagnosis did not decline during the pandemic and even increased among those aged 80 and over (Fig SXXA).
The decrease in cancer-attributed deaths was concentrated among cancers with low baseline survival (Fig SXXB).
Conversely, among individuals with cancer whose underlying cause of death was not cancer, the largest excess was observed in those with the least lethal cancer types (Fig SXXC).

## Correlation with respiratory diseases in 2020-21

We examined the temporal correlation between excess mortality from specific causes and excess mortality due to either COVID-19 or other respiratory diseases during 2020–21.
For cardiovascular and mental/neurological diseases, we observed positive partial correlations with COVID-19 mortality, reaching nearly 50% among individuals aged 80 and over (Figure 5).
Similar correlations were also found with other respiratory diseases, reflecting the high correlations inferred by the model for the prepandemic period 2011-2019.
The strongest correlations with COVID-19 were seen at lags of 1 to 3 weeks, meaning that COVID-19-related excess deaths happened around the same time as, but slightly after, those from cardiovascular or mental/neurological causes.
In contrast, we observed slightly negative correlations between cancer and COVID-19 mortality, particularly among those aged 65–79.

-> undetected rather than long-term consequence of covid, as peaks occur simultaneously






```{r, include=FALSE}

```



Cancer : 
Reclassification of deaths among cancer patients with COVID-19: Many cancer patients who contracted COVID-19 during the 2nd and 3rd waves likely died with contributions from both diseases, but their deaths were primarily attributed to COVID-19 (as per guidelines), reducing recorded cancer mortality. This effect was stronger in 65–79, who were more likely to be hospitalized and tested.
+ harvesting effect of cancer patients occured in wave 2 lead to decrease in deaths in wave 3 
+ absences of resp infections during wave 3
  
```{r, include=FALSE}

```







Reasons: 
1) Undiagnosed COVID-19 case : -> increase during wave
2) Long-term mortality displacement -> decrease after waves
3) Absence of resp -> decrease in Jan/Feb
4) Direct consequence of COVID-19: -> increase during wave
5) Short-term mortality displacement -> decrease during wave

Different to say the same
1) Wrong ascertainment : considered as condition (low proportion of condition, among covid deaths) -> increase
5) Wrong ascertainment : considered as covid (high proportion of condition, e.g., cancer, among covid deaths) -> decrease

during wave: + 1), + 4), - 5)
after wave: - 2), + 4) (as a long-term consequence of covid, but not really observed in the results)
Winter: - 3)


among 65-79: no correlation during covid between resp (or covid) and mental (as we observe before pandemic), likely because ascertainment of covid deaths changed over time  (e.g., through testing strategies), as shown by time-varying proportion of covid deaths with mental diseases$



TO DO:
- mention that excess deaths before 2020 coincides with episode of flu epidemic -> undiagnosed or consequence of infection
- look at overall mortality by cause accounting for principal cause, direct cause and underlying condition.
Compare between covid and pre-covid, especially for cancer in order to see whether overall cancer mortality really decreased or whether it's only because of different attribution of cause of deaths.
could inspire from : https://elifesciences.org/reviewed-preprints/93758?utm_source=chatgpt.com


modest attribution shift of covid-cancer deaths

\newpage

# Discussion

## Summary

We modeled cause-specific mortality trends in Switzerland (2011–2019) to assess excess mortality during the COVID-19 pandemic (2020–2021).
Compared to the pre-pandemic period, we observed several disruptions in mortality patterns.
Cardiovascular deaths increased temporarily during COVID-19 waves, particularly among those aged 65 and older.
In contrast, respiratory disease mortality declined consistently throughout the pandemic.
For mental/neurological, and cancer-related causes in older adults, mortality deficits were mostly limited to 2021.
These deficits were offset by COVID-19 deaths, resulting in persistent all-cause excess mortality.
These shifts likely reflect a combination of mechanisms—such as mortality displacement, competing risks and reduced circulation of other pathogens due to control measures—that are explored in more detail below.
<!-- synchrony and correlation of excess deaths, observed before covid and after -->

## Potential explanations for the observed patterns: 4 types of disruption



There are four types of disruptions that could serve as potential explanations for the observed patterns.
First, the under-ascertainment of COVID-19 mortality lead to an artificial inflation of the deaths due to other causes.
In Switzerland, Riou et al. estimated that only 72% of COVID-19 related deaths were reported as such in 2020-21, with even lower proportion during epidemic waves.
As SARS-CoV-2 can contribute to cardiovascular death through pathways involving inflammation, thrombosis, and direct cardiac injury, misclassified COVID-19 deaths might have been considered as due to cardiovascular causes.
This might indeed explain the excess mortalities due to cardiovascular diseases observed during both covid wave in fall 2020 and fall 2021.
<!-- ascertainment of cases: Bouman -->
<!-- Riou: only 72% of COVID-19 were reported, modelling work, even lower among 80+, where we estimated more pronounced fluctuation of excess mortality during covid waves. -->

A second type of disruption concerns the sequelae following SARS-CoV-2 infection that might affect cause-specific mortality.
Several studies have emphasized the increased risk of developing a certain number of sequela or symptoms after SARS-CoV-2 infections.
This could lead to an excess mortality, as found by Cai et al. in the US.
For example, as this sequelas often affects the cardiovascular system, we might experience excess mortality due to cardiovascular diseases after a covid-19 wave.
This effect occurs with delays and is potentially long-lasting, is quite limited, so that it is might be hard to identify in the Swiss population, which is healthier than the US.
It might as well be concurrent with other type of disruption, which might blurr the interpretation of the results.

Third, mortality displacement, i.e., when deaths occur earlier than they would have in the absence of the COVID-19 pandemic, partly explains the deficit in mortality observed during the first part of 2021.
This was mostly apparent for mental/neurological diseases, where we observe a strong deficit of mortality of 65+ during the period following COVID-19 waves in Fall 2020 and winter 2020-21.
Such undermortality has already been observed in other countries, e.g., in Nordic countries, attributing it to displacement from dementia deaths to COVID-19 deaths.

While mortality displacement might affect mortality over year, it has a more immediate effect on causes with low life prognosis, such as certain type of cancers.
Additional analyses showed that the apparent decrease in cancer mortality observed end of 2020 was compensated by an increase in the number of individuals with cancer dying from other causes (such as COVID-19).
It also showed that this decrease was mostly visible in cancers with low prognosis, where competing risk was highest.
<!-- ADD sentence to justify this in more details, citing the paper about modelling competing risk -->
<!-- ADD harvesting -->
The deficit in mortality observed almost simultaneously as the covid-19 wave among 65-79 could therefore be explained by mortality displacement over causes.
This also explains the negative correlation calculated between COVID-19 deaths and excess mortality due to cancer.
<!-- cite paper us about correlation over states -->

Finally, the pandemic response measures, whether mandated through control measures or voluntary behavioural changes, have affected mortality in various ways.
Non-pharmaceutical interventions (NPIs) have reduced the circulation of many respiratory pathogens (e.g., flu), leading to a deficit in mortality caused by respiratory diseases.
<!-- mask wearing, social distancing  -->
This deficit is stricking in January-February 2021, as it is a period of the years usually marked by intense circulation of respiratory pathogens but not in 2021 where strict control measures were implemented.
In addition to the direct effect on respiratory disease mortality, the absence of circulation of respiratory pathogens is likely to have affected other causes of deaths, especially among 80+, e.g., due to the dependencies with neurological/mental or cardiovascular diseases.
This could partly explain the deficit in mortality due to mental/neurological diseases observed in the beginning of 2021 in the 65-79 and 80+ age groups.
Alternatively, the restrictions imposed during COVID-19 had some repercussions on the standard of care.
This has been used to justify the lower number of cancer deaths as a result of the underdiagnosis.
However, Switzerland, by implementing moderate control measures, was able to keep a high standard of care throughout the pandemic, making this explanation unrealistic.

The multivariate structure of our model identified two levels of correlation/association between cause-specific mortality before 2020.
First, most of the causes have mortality peak at the same time during winter in multiple age groups.
But our model also shows that, for respiratory, cardiovascular and mental/neurological diseases, the excesses remain strongly correlated after adjusting for seasonal variation.
The correlation is highest in winters with the largest influenza epidemics, implying that a substantial share of the apparent cardiovascular excess is actually infection-related (either biologically triggered events or misclassified deaths).
While extreme weather or pollution could still contribute, previous studies indicate it explains only a minor fraction of the residual synchrony that we observe.
This is validated by the correlations between either cardiovascular or mental/neurological and both covid-19 and other respiratory excesses during 2020-21, a period where traditional mortality patterns were disrupted.
In this context, covid disruption serves as a natural experiment, as covid wave occurred at unusual periods for a respiratory disease and flu wave was almost inexistant in winter 2020-21.
While the mechanisms still need to be investigated in more depth, this strongly suggests that the impact of respiratory pathogens on mortality are likely underestimated.

- multivariate (cross-cause random effect) + seasonality + long-term trend + population change
- posterior propagation of uncertainty
- complete coverage of deaths, while we use underlying deaths, other variables could be used for in-depth investigation (e.g., cancer)

- mortality displacement: needs further investigation but difficult with only deaths data
  - competing risk not properly handled: mortality displacement
  - frailty model : account for heterogeneity inside group
- misclassification (as we use underlying deaths):
  - we acknowledge and try to interpret
  - we run in depth investigation using other variable to look at this impact
- broad age groups; static regional and sex effect
- ecological level: difficult to interpret correlations
- analysis stops at 2021: no long-term effect

- apply same approach to other countries to validate correlation
- investigate this by socio-economic status: level of vulnerabilities
- more recent years

cause of death data analysis is important because:
- understanding mechanisms and depedencies
- predict all-cause mortality, Huynh
- explain differences observed by countries


model: able to correct for trend while propagating uncertainty which is key as excess mortality is about prediction and conterfactuals,
accounting for population change and for correlation between cause
data: accurate information, with clear and consistent over time guidelines to code deaths, and with other variables available to dig about conditions (e.g., for cancer, competing risks)
weekly data -> granulate temporal information


main cause of deaths: difficult to assess in some deaths with multiple comorbidities
difficult to interpret numbers in young population (suicide) as deaths concentrate in older
same for effect of regions
aggregation by cause is arbitrary
separate by age/aggregation by age (i.e, ppl within same age group have identical risk)

competing risk models might be useful to account for mortality displacement over causes




1) Undetected COVID-19 deaths
synchrony observed between cardiovascular excess and covid waves
-> undetected covid, attributed to cvd, makes senses as covid directly alters cardiovascular system
(-> could be also consequence of covid but less likely as occurs same time)
mental/neurological among 80+

2) Effect of covid-19 infection (indistinguishable from undetected covid-19 deaths if the effect is short-term)
Cai et al.
post-acute sequelae condition with CVD symptoms higher after infection, mental/neurological incidence rate  higher, short-term
also higher all-cause excess deaths after adjusting for potential confounders

Fares
Warren-Gash
Han

3) Deficit in mortality in 2021 can be explained by also mortality displacement
long-term: cvd, dementia (harvesting), died from covid instead of cancer
shorter term: cancer (due to competive risks), die
cvd in 2021, dementia
peak in mortality observed in autumn 2020 (either due to covid-19 or other causes) means that some of the deaths would have occurred anyway, harvesting

competing risk in cancer -> more impact among 65-79, where mortality rate between cancer and covid are more similar

4) Impact of control measures (indirect impact of covid)
4b) Care disruption: unlikely, as shown by other study for cancer
or underdiagnosis -> leading to undetected cancer deaths
4a) Specific deficit in Jan-Feb explained by resp absence (short term)
correlated with respiratory, suggesting an effect
Hygiene measure: absence of mortality peak due to respiratory disease.



Correlation

Two levels : 1) Mortality, 2) Excess mortality, exceptional event
first level could be due to confounders, shared drivers (cold, winter -> confinement inside)
second level makes it more difficult to believe confounder, as they stay constant

some mechanisms (undetected covid-19 deaths, short-term effect) also explains the correlation of cause of deaths during covid
positive for cvd and 
negative for cancer: also observed by Luck
same with other respiratory diseases (maybe don't mention it because difficult to explain)
same as in prepandemic, high correlation with resp mortality (cvd and mental/neurological diseases): 2 levels of correlation (seasonality and excess)
synchrony excess deaths
-> natural experiment during covid (link btw resp and cvd/)
mechanisms/pathways: Fares, Warren-Gash




## Comparison with other studies

1) Nilsson, excess mortality by cause in scandinavia
Title: Cause-specific excess mortality in Denmark, Finland, Norway, and Sweden during the COVID-19 pandemic 2020–2022: a study using nationwide population data
model: linear regression
results: excess mortality for cvd

1) C‑MOR Consortium (~12 countries): Beeks, Achilleos
Title: Cause-Specific Excess Mortality During the COVID-19 Pandemic (2020–2021) in 12 Countries of the C-MOR Consortium
model: two methods, sinusoidal curve to model seasonality or 5 years before pandemic as baseline (Beeks or Achilleos)
multilevel (mixed-effects)
results: conflicting results (btw countries) for cancer, increased mortality for cvd
positive association between covid incidence and cvd excess
negative association between stringency of control measures and cvd (in 2021)

2) Paglino, SOA
Title: Cause-Specific Excess Mortality During the COVID-19 Pandemic
model: SARIMA often underestimates long-term uncertainty
results: strong correlation between covid-19 deaths and excess from cvd
but it's US, so difficult to compare, still interesting to see that long-term effect (scarring effect) exists there in contrast to CH.
comment: GP more suitable with sudden deviations from seasonal norms

3) Zhang: in Italy
Title: Bayesian Poisson Regression and Tensor Train Decomposition Model for Learning Mortality Pattern Changes during COVID-19 Pandemic
model: tensor train decomposition, captures latent structure and dependencies in the high dimensional resiudals
regional temporal clustering, cause of deaths patterns
comment: mostly used to identify similar temporal or geographical patterns, difficult to identify association between causes

4) Han:
Title: Excess cardiovascular mortality across multiple COVID-19 waves in the United States from March 2020 to March 2022
model: log-linear model with linear temporal trend and periodic spline
results: excess mortality for cvd, due to cardiovascular complication after sars-cov-2 infection and long-term elevated cvd risk, and to reduced cvd care services.

5) Huynh: 
Title: Joint models for cause-of-death mortality in multiple populations
model: joint modelling age-specific rates of various causes of deaths, Multi-Output Gaussian Processes (MOGP)
results: cannot use them, show how all-cause forecasts benefit from modeling cause interdependencies
comments: very theoretical, no results

8) Pizzato: correlation between covid and cause-specific excess mortality in 16 countries 2020-23.
Title: Relationship between COVID-19 cases and monthly mortality from all causes, cancer, cardiovascular diseases and diabetes in 16 countries, 2020–21
model: generalized linear regression with linear trend and seasonal trend with splines, spline also for age
results: positive correlation btw cvd and covid in most countries, no excess for cancer (sometimes lower than expected)

6) Luck: age-standardize deaths
Title: Associations between mortality from COVID-19 and other causes: A state-level analysis
model: correlation across states between cause of deaths with covid
results: negative correlation for cancer

7) Carollo
Title: Competing risks models with two time scales
comments: only interesting for competing risk not for cause correlation


## Natural experiment: synchrony between flu and CV/dementia

## Potential explanations for the observed patterns: 3 types of disruption

## Strengths, limitations, perspectives

model: able to correct for trend while propagating uncertainty which is key as excess mortality is about prediction and conterfactuals,
accounting for population change and for correlation between cause
data: accurate information, with clear and consistent over time guidelines to code deaths, and with other variables available to dig about conditions (e.g., for cancer, competing risks)
weekly data -> granulate temporal information


main cause of deaths: difficult to assess in some deaths with multiple comorbidities
difficult to interpret numbers in young population (suicide) as deaths concentrate in older
same for effect of regions
aggregation by cause is arbitrary
separate by age/aggregation by age (i.e, ppl within same age group have identical risk)

competing risk models might be useful to account for mortality displacement over causes


## Conclusion
  
  
  
  \newpage

# Figures

```{r, fig.width=10,fig.height=10}
fig1 = res_list$Sigma_mat %>% 
  dplyr::mutate(cod_group = factor(cod_group,levels=causes2_df$cod_group,
                                   labels=causes2_df$cod_group_label),
                cod_group2 = factor(cod_group2,levels=causes2_df$cod_group,
                                    labels=causes2_df$cod_group_label),
                cod_group_id=as.numeric(cod_group),
                cod_group_id2=as.numeric(cod_group2)) %>% 
  rowwise() %>% 
  dplyr::mutate(est_cri = paste0(scales::percent(est, accuracy = 1),"\n",
                                 "[",scales::percent(lwb, accuracy = 1),",",
                                 scales::percent(upb, accuracy = 1),"]")) %>% 
  filter(cod_group_id<cod_group_id2,age_class=="80+") %>% 
  ggplot(aes(x = cod_group, y = fct_rev(cod_group2), fill = abs(est))) +
  geom_tile() +
  geom_text(aes(label = est_cri),
            color = "black", size = 3) +
  scale_fill_gradient(low = "lightyellow", high = "orangered1",limits=c(0,1),
                      name="Correlation",
                      labels=scales::label_percent(accuracy = 1)) +
  labs(x = "", y = "", fill = "Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0,size=11),
        axis.text.y = element_text(size=11))+
  scale_x_discrete(position = "top")
fig1
```

```{r, include=FALSE}
pdf(file="fig1.pdf",width=10,height=10)
print(fig1)
dev.off()
```

**Figure 1.** Social contact data and mobility data.
(A) Number of participants by survey wave and by country. The opaque part corresponds to the newly enrolled participations (i.e., contributing for the first time), while the more transparent part corresponds to participants that have already taken part in the survey before.
(B) Age distribution. The colored bars display the number of participants by age classes for the four countries. The grey bars in the backgrounds reflect the age distribution in the general population.
(C) Distribution of the number of survey round participated.
(D) Google work-related mobility data. The mobility levels during the weekdays (Monday-Friday) and weekends (Saturday-Sunday) are averaged weekly.


```{r, fig.width=8,fig.height=6}
fig2=fig1
```

```{r, include=FALSE}
pdf(file="fig2.pdf",width=8,height=6)
print(fig2)
dev.off()
```

**Figure 2.** Survey fatigue and effect of mobility on contacts.
(A) Survey fatigue on the probability of reporting no contact $\theta$. The odds ratio (OR) of reporting zero contact is displayed according to the survey round and country
(B) Survey fatigue on the contact rate $\lambda$. The relative number of contacts (relatively to the contact rate $\lambda$) is displayed according to the survey round and country.
(C) Effect of mobility on the probability of reporting no contact and on the contact rate. The estimated effects are displayed for the four countries and with or without adjusting for long-term trend through Gaussian process (GP).

```{r, fig.width=9,fig.height=12}
fig3=fig1
```

```{r, include=FALSE}
pdf(file="fig3.pdf",width=9,height=12)
print(fig3)
dev.off()
```

**Figure 3.** Model fits for the UK.
(A) Probability of having contacts at work, as estimated by the survey model (red points with vertical lines) and as predicted by the mobility model (colored lines). We shows the predictions of three versions of the mobility model: 1) considering the full period (black), 2) adjusting for the long-term change of contact rate (blue) and 3) limiting to the period until February 2021 (green). For the probability of having contacts at work, we do not show the results of the full period without adjustment of the long-term change, as it is almost indistinguishable from the analysis when adjusting for it. The red points corresponds to the mean posterior estimates by survey wave, and the vertical line shows the related 95% credibility interval (95%CrI).
The colored lines correspond to the (weekly) mean posterior estimates and the gray area corresponds to the 95%CrI. The exact definition of the posterior estimates is given in Supplementary S1.3.2.
(B) Contact rate at work as estimated by the survey model (red) and as predicted by the mobility model (black).
(C) Probability of having contacts at work as estimated by the survey model (red) and as predicted by the mobility model (black), according to mobility level. The different levels of darkness in the red estimates of the survey model depicts the different survey waves.


```{r, fig.width=8,fig.height=6}
fig4 = fig1
```

```{r, include=FALSE}
pdf(file="fig4.pdf",width=8,height=6)
print(fig4)
dev.off()
```

**Figure 4.** Baseline estimates.
(A) Baseline proportion of individuals having work-related contacts as estimated by various prepandemic data and by the mobility model, stratified by age classes.
The points corresponds to the mean estimate and the vertical bars to the 95%CrI.
(B) Baseline conditional number of work-related contacts as estimated by various prepandemic data and by the mobility model, stratified by age classes.

\newpage

# Data sharing statement

# Acknowledgements

\newpage

# References


