---
title: "COVID-19 disrupted patterns of cause-specific mortality in Switzerland"
date: "`r format(Sys.time(), 'Report generated on Date: %Y-%m-%d Time: %H:%M')`"
author: "Anthony Hauser* (1), Moritz Wagner (2), Anne Laube (2), Julien Riou (1)"
output:
    word_document: 
      reference_docx: template.docx
      number_sections: true
      df_print: paged
bibliography: bibliography.bib
csl: ieee.csl
---

(1) Department of Epidemiology and Health Systems, Unisanté, Center for Primary Care and Public Health and University of Lausanne, Lausanne, Switzerland;<w:br/>
(2) Federal Office of Public Health, Bern, Switzerland;<w:br/>
(*) Corresponding author: anthony.hauser@unisante.ch

```{r setup, include=TRUE}
#Setup
Sys.setlocale("LC_TIME", "C")
wd = getwd()
code_root_path = paste0(strsplit(wd, split="/manuscript")[[1]][1],"/")
print(code_root_path)
source(paste0(code_root_path,"R/000_setup.R"))
knitr::opts_chunk$set(echo = FALSE, warning =FALSE, message=FALSE)

wd_ofsp = "L:/UNISANTE_DESS/S_SUMAC/OFSP_2023/"
data_folder = paste0(wd_ofsp,"02_data/cause_of_death/")
save.date="20241218"
```


```{r, include= TRUE}
#Individual dataset
cod_ind_df = readRDS(paste0(code_root_path,"savepoint/cod_ind_df.RDS"))

causes2 = c("Cardiovascular Diseases","Infectious and Parasitic Diseases",
           "Respiratory Diseases", "Mental and Neurological Disorders",
           "COVID-19",
           "Neoplasms (Cancers)","Suicide","External Causes")

causes2_df = data.frame(cod_group=c("Cardiovascular Diseases","Infectious and Parasitic Diseases",
                                    "Respiratory Diseases", "Mental and Neurological Disorders",
                                    "COVID-19",
                                    "Neoplasms (Cancers)","Suicide","External Causes","Other Causes"),
                        cod_group_label=c("Cardiovascular","Infectious/Parasitic",
                                    "Respiratory", "Mental/Neurological",
                                    "COVID-19",
                                    "Cancers","Suicide","External","Other"),
                        order=c(1,7,4,3,9,2,6,5,8),
                        example = c("Ischaemic heart disease, heart attack", "Sepsis",
                                   "Pulmonary disease, pneumonia, influenza", "Dementia, Alzheimer, Parkinson",
                                   "COVID-19",
                                   "Lung, breast, prostate neoplasms","Intentional self-harm/poisoning","Fall, vehicule accident",
                                   "Unspecified, diabetes, senility, organ disease")) %>%
  #add chapter of cod-10
  left_join(cod_ind_df %>%
              filter(outcome=="ENDG_U_CD_GES_T") %>%
              dplyr::select(cod_group,chapter) %>% unique() %>%
              arrange(cod_group,chapter) %>%
              group_by(cod_group) %>%
              dplyr::summarise(chapter=paste(chapter, collapse=", ")),by="cod_group") %>%
  arrange(order)
```

```{r, include=FALSE}
#Aggregated dataset and results
add_apo = function(x){format(x, big.mark=",")}

cod_agg_pop_df = readRDS(paste0(code_root_path,"savepoint/cod_agg_pop_df.RDS"))
age_classes = cod_agg_pop_df$age_class %>% unique()
setwd(code_root_path)
res_list = load_results_mod6(age_classes, save.date=save.date,mod="mod8")

cod_agg_pop_df = cod_agg_pop_df %>%
  dplyr::mutate(date=ISOweek2date(paste0(cal_year,"-W",ifelse(cal_week<10,paste0("0",cal_week),cal_week),"-1"))) %>% 
  mutate(covid_phase = map2_dbl(date, list(covid_phase), function(d, phases) {
    phase <- phases %>%
      filter(d >= start_date & d <= end_date) %>%
      pull(phase)
    if (length(phase) == 0) NA_real_ else phase
  }))
```


# Abstract

```{r, include=FALSE}

```

**Introduction**
  
  
**Methods and Findings**
  
  
**Conclusion**
  
  
\newpage

# Introduction

Link between covid and other causes
direct: 
indirect: measures -> suicides, respiratory diseases, accident

-> correlation of mortality causes
  -> important to analyse them simultaneously (as increase of a cause can lead to decrease of another cause, and thus no all-cause excess)
  
Disruption of mortality patterns offers a unique opportunity to explore the nature/origin of correlation between causes (which almost all usually peaks during winter but why? is it due to the conditions? to resp virus? other?)

Gap: cause-specific mortality during covid


Objectives

\newpage

# Methods

## Data

We received individual-level data on cause-specific deaths for the years 2011–2021 from the Federal Statistical Office (FSO), which we aggregated by age group (0–17, 18–39, 40–64, 65–79, 80+), sex, NUTS-2 region, calendar week, and calendar year.
The data include information recorded in the death certificate—such as the immediate, underlying, and contributing causes of death—as well as the official underlying cause of death, which is determined by the FSO based on the coding rules of the International Classification of Diseases, 10th Revision (ICD-10).
All causes of death are coded using ICD-10 and were grouped into nine categories for analysis: Cardiovascular diseases (Chapter IX), Cancers (Chapter II), Mental and neurological disorders (Chapters V and VI), Respiratory diseases (Chapter X), External causes (Chapter XX), Suicide (Chapter XX), Infectious and parasitic diseases (Chapter I), Other causes (Chapters III, IV, VII, VIII, XI–XIX, XXI, XXII), and COVID-19 (Chapter XXII).
Population data by sex, NUTS-2 region, and age group for the years 2011–2021 were also provided by the Federal Statistical Office (FSO). We applied linear interpolation to obtain weekly population estimates.

<!-- In Switzerland, the internal cause-specific mortality database records detailed information on the circumstances of each death and the characteristics of the deceased. -->
<!-- For every death of a Swiss resident, a physician completes a death certificate specifying the immediate, underlying, and contributing causes of death. -->
<!-- The death certificate is then transmitted to the Federal Statistical Office (FSO), where information is anonymized, coded, and entered into the national mortality database -->
<!-- The official underlying cause of death is determined according to the coding rules of the International Classification of Diseases, 10th Revision (ICD-10). -->
<!-- Additional variables — such as age, canton of residence, calendar week, and year of death — are also recorded. -->


## Model

<!-- We modeled the weekly number of deaths by cause for each age group using a Poisson distribution with a multivariate structure.   -->
<!-- The model accounts for changes in population size $N$ over time via an offset term, and includes fixed effects $\beta$ and $\gamma$ for differences in mortality risk by sex and region, respectively. -->
<!-- Temporal variation is captured by two components: a seasonal trend modeled with a periodic Gaussian Process (GP) $f(t)$, and a long-term trend modeled using a GP with an exponentiated quadratic (squared exponential) kernel $g(t)$.   -->
<!-- To account for the correlation between mortality risks from different causes, the model incorporates a multivariate random effect $\epsilon$. -->
<!-- The model reads as follow: -->
<!-- $$y_{i}(t)\sim Poisson(\mu_{i}(t))$$ -->
<!-- with, -->
<!-- $$\log\mu_{irs}(t) = \alpha_i + \beta_{ir} + \gamma_{is} + f(t) + g(t) + \epsilon_i + \log N_{irst}$$ -->
<!-- where $y_{irs}(t) $ is the observed number of deaths due to cause $i$ at week $t$ in region $r$ and sex $s$. -->
<!-- The parameters $\alpha_i$ represents the baseline relative mortality risk. -->

We modeled the weekly number of deaths by cause for each age group using a Poisson distribution with a multivariate structure.  
The model accounts for changes in population size $ N $ over time via an offset term, and includes fixed effects $ \beta $ and $ \gamma $ for differences in mortality risk by region and sex, respectively.  
Temporal variation is captured by two components: a seasonal trend modeled with a periodic Gaussian Process (GP) $ f(t) $, and a long-term trend modeled using a GP with an exponentiated quadratic (squared exponential) kernel $ g(t) $.  
To account for the correlation between mortality risks across causes, the model incorporates a multivariate random effect $ \epsilon \sim \mathcal{N}(0,\Sigma)$, where $\Sigma$ is the correlation matrix 

The model is specified as follows:
$$
y_i(t) \sim \text{Poisson}(\mu_i(t))
$$
with
$$
\log \mu_{irs}(t) = \alpha_i + \beta_{ir} + \gamma_{is} + f(t) + g(t) + \epsilon_i + \log N_{irst},
$$
where $ y_{irs}(t)$ denotes the observed number of deaths due to cause \( i \) in week \( t \), region \( r \), and sex \( s \), and \( N_{irst} \) is the corresponding population at risk.
<!-- where $ y_{irs}(t) $ denotes the observed number of deaths due to cause $ i $ at week $ t $, in region $ r $ and sex $ s $; $ \alpha_i $ is the baseline log mortality risk for cause $ i $; $ \beta_{ir} $ and $ \gamma_{is} $ are age-specific fixed effects for region and sex, respectively; $ f(t) $ captures the seasonal variation; $ g(t) $ represents the long-term trend; $ \epsilon \sim \mathcal{N}(0,\Sigma)$ is a multivariate random effect capturing correlation across causes (with correlation matrix $\Sigma$); and $ N_{irst} $ is the population at risk, included as a log offset. -->


The model was run independently for each age group and fitted to the weekly number of cause-specific deaths $y_{irs}(t)$ observed between 2011 and 2019.
It was then used to estimate the weekly \textit{expected number of deaths}  $\tilde{y}_{irs}(t)$ through the end of 2021, defined as draws from the \textit{posterior predictive distribution} after removing the random effect $\epsilon$ from the log mortality risk $\mu_{irs}(t)$.  
We excluded $\epsilon$ because it is designed to capture \textit{excess mortality}—that is, unexplained or anomalous variation beyond structured trends.  
Excess mortality $E_{irs}(t)$ was then defined as the difference between observed and expected deaths:
$$E_{irs}(t) := y_{irs}(t) - \tilde{y}_{irs}(t).$$
where $y_{irs}(t)$ is the observed number of deaths and $\tilde{y}_{irs}(t)$ the predicted number under baseline conditions.  
We also report aggregated estimates of excess mortality, such as \textit{all-cause excess mortality} (aggregated over causes of death) or aggregated over time period.

## Further analysis



Correlation
Correlation during COVID-19: investigate the partial correlation between cause-specific excess mortality rates during covid-19 between different causes and respiratory disease, which 


Q: where to describe covid-19 phase in switzerland? Introduction?
\newpage

# Results

## Distribution of deaths

In Switzerland, deaths are heavily concentrated among older adults: 88% occurred in individuals aged 65 and over, while fewer than 2% were among those under 40 in 2020-21 (Fig 1A).
Cardiovascular diseases were the leading cause of death overall, followed by cancers and mental or neurological disorders, though the distribution varied by age group  (Fig 1B).
Among individuals aged 80 and above, cardiovascular disease accounted for XX% of all deaths.
In contrast, for those aged 40–79, cancer was the predominant cause, responsible for approximately 40% of deaths.
The relative impact of COVID-19 was greatest in older age groups, contributing to XX% of deaths among those aged 65–79 and XX% among those aged 80 and over.

## Prepandemic mortality trends

The model successfully captures both the seasonal and long-term trends in cause-specific mortality observed between 2011 and 2019 (Figure 2A for individuals aged 80+, and Figures SXX–SXX for other age groups).
We observe notable similarities in mortality trends across different causes, both in the expected seasonal patterns and in intermittent episodes of excess mortality that often occur synchronously (e.g., during the winters of 2014/15 and 2016/17).
For the 80+ age group, Figure 2B highlights this synchrony in timing, showing that the expected mean mortality peak (in black) occurs consistently across causes between January and February.
During the pre-pandemic period (2011–2019), observed peak timings (colored circles) align closely with these expectations.
However, during the pandemic years, we observe a marked shift: mortality peaks occur earlier than expected (colored squares).
In addition, the model reveals strong positive correlations in cause-specific excess mortality, mostly among individuals aged 80 and over (Fig 2C).
Specifically, excess deaths from respiratory diseases, mental or neurological disorders, and cardiovascular diseases are highly correlated, with pairwise correlations exceeding 80%.

## COVID-19 mortality

The model also captures all-cause mortality, obtained by aggregating deaths across causes.
Among individuals aged 65 and over, the impact of the first two COVID-19 waves—-spring 2020 and fall/winter 2020–21—-clearly exceeded any episodes of excess mortality observed between 2011 and 2019 (Figure 3A).
Cumulative excess mortality highlights the substantial increase in deaths during these waves, followed by a gradual mortality deficit in the subsequent weeks (Figure 3B).
When excluding deaths attributed to COVID-19, we also observe that mortality from other causes was lower than expected in 2020–21 for people aged 65 and over.




```{r, include=FALSE}

```



Cancer : 
Reclassification of deaths among cancer patients with COVID-19: Many cancer patients who contracted COVID-19 during the 2nd and 3rd waves likely died with contributions from both diseases, but their deaths were primarily attributed to COVID-19 (as per guidelines), reducing recorded cancer mortality. This effect was stronger in 65–79, who were more likely to be hospitalized and tested.
+ harvesting effect of cancer patients occured in wave 2 lead to decrease in deaths in wave 3 
+ absences of resp infections during wave 3

## Excess all-cause
  
## CV, cancer, dementia and infectious or respiratory diseases
  
```{r, include=FALSE}

```



Reasons: 
1) Undiagnosed COVID-19 case : -> increase during wave
2) Long-term mortality displacement -> decrease after waves
3) Absence of resp -> decrease in Jan/Feb
4) Direct consequence of COVID-19: -> increase during wave
5) Short-term mortality displacement -> decrease during wave

Different to say the same
1) Wrong ascertainment : considered as condition (low proportion of condition, among covid deaths) -> increase
5) Wrong ascertainment : considered as covid (high proportion of condition, e.g., cancer, among covid deaths) -> decrease

during wave: + 1), + 4), - 5)
after wave: - 2), + 4) (as a long-term consequence of covid, but not really observed in the results)
Winter: - 3)


among 65-79: no correlation during covid between resp (or covid) and mental (as we observe before pandemic), likely because ascertainment of covid deaths changed over time  (e.g., through testing strategies), as shown by time-varying proportion of covid deaths with mental diseases$



TO DO:
look at overall mortality by cause accounting for principal cause, direct cause and underlying condition.
Compare between covid and pre-covid, especially for cancer in order to see whether overall cancer mortality really decreased or whether it's only because of different attribution of cause of deaths.
could inspire from : https://elifesciences.org/reviewed-preprints/93758?utm_source=chatgpt.com


modest attribution shift of covid-cancer deaths

\newpage

# Discussion

**Summary**
  
**Comparison with other studies**

**Natural experiment: synchronicity between flu and CV/dementia**

**Potential explanations for the observed patterns: 3 types of disruption**

**Strengths, limitations, perspectives**

**Conclusion**
  
  
  
  \newpage

# Figures

```{r, fig.width=10,fig.height=10}
fig1 = res_list$Sigma_mat %>% 
  dplyr::mutate(cod_group = factor(cod_group,levels=causes2_df$cod_group,
                                   labels=causes2_df$cod_group_label),
                cod_group2 = factor(cod_group2,levels=causes2_df$cod_group,
                                    labels=causes2_df$cod_group_label),
                cod_group_id=as.numeric(cod_group),
                cod_group_id2=as.numeric(cod_group2)) %>% 
  rowwise() %>% 
  dplyr::mutate(est_cri = paste0(scales::percent(est, accuracy = 1),"\n",
                                 "[",scales::percent(lwb, accuracy = 1),",",
                                 scales::percent(upb, accuracy = 1),"]")) %>% 
  filter(cod_group_id<cod_group_id2,age_class=="80+") %>% 
  ggplot(aes(x = cod_group, y = fct_rev(cod_group2), fill = abs(est))) +
  geom_tile() +
  geom_text(aes(label = est_cri),
            color = "black", size = 3) +
  scale_fill_gradient(low = "lightyellow", high = "orangered1",limits=c(0,1),
                      name="Correlation",
                      labels=scales::label_percent(accuracy = 1)) +
  labs(x = "", y = "", fill = "Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0,size=11),
        axis.text.y = element_text(size=11))+
  scale_x_discrete(position = "top")
fig1
```

```{r, include=FALSE}
pdf(file="fig1.pdf",width=10,height=10)
print(fig1)
dev.off()
```

**Figure 1.** Social contact data and mobility data.
(A) Number of participants by survey wave and by country. The opaque part corresponds to the newly enrolled participations (i.e., contributing for the first time), while the more transparent part corresponds to participants that have already taken part in the survey before.
(B) Age distribution. The colored bars display the number of participants by age classes for the four countries. The grey bars in the backgrounds reflect the age distribution in the general population.
(C) Distribution of the number of survey round participated.
(D) Google work-related mobility data. The mobility levels during the weekdays (Monday-Friday) and weekends (Saturday-Sunday) are averaged weekly.


```{r, fig.width=8,fig.height=6}
fig2=fig1
```

```{r, include=FALSE}
pdf(file="fig2.pdf",width=8,height=6)
print(fig2)
dev.off()
```

**Figure 2.** Survey fatigue and effect of mobility on contacts.
(A) Survey fatigue on the probability of reporting no contact $\theta$. The odds ratio (OR) of reporting zero contact is displayed according to the survey round and country
(B) Survey fatigue on the contact rate $\lambda$. The relative number of contacts (relatively to the contact rate $\lambda$) is displayed according to the survey round and country.
(C) Effect of mobility on the probability of reporting no contact and on the contact rate. The estimated effects are displayed for the four countries and with or without adjusting for long-term trend through Gaussian process (GP).

```{r, fig.width=9,fig.height=12}
fig3=fig1
```

```{r, include=FALSE}
pdf(file="fig3.pdf",width=9,height=12)
print(fig3)
dev.off()
```

**Figure 3.** Model fits for the UK.
(A) Probability of having contacts at work, as estimated by the survey model (red points with vertical lines) and as predicted by the mobility model (colored lines). We shows the predictions of three versions of the mobility model: 1) considering the full period (black), 2) adjusting for the long-term change of contact rate (blue) and 3) limiting to the period until February 2021 (green). For the probability of having contacts at work, we do not show the results of the full period without adjustment of the long-term change, as it is almost indistinguishable from the analysis when adjusting for it. The red points corresponds to the mean posterior estimates by survey wave, and the vertical line shows the related 95% credibility interval (95%CrI).
The colored lines correspond to the (weekly) mean posterior estimates and the gray area corresponds to the 95%CrI. The exact definition of the posterior estimates is given in Supplementary S1.3.2.
(B) Contact rate at work as estimated by the survey model (red) and as predicted by the mobility model (black).
(C) Probability of having contacts at work as estimated by the survey model (red) and as predicted by the mobility model (black), according to mobility level. The different levels of darkness in the red estimates of the survey model depicts the different survey waves.


```{r, fig.width=8,fig.height=6}
fig4 = fig1
```

```{r, include=FALSE}
pdf(file="fig4.pdf",width=8,height=6)
print(fig4)
dev.off()
```

**Figure 4.** Baseline estimates.
(A) Baseline proportion of individuals having work-related contacts as estimated by various prepandemic data and by the mobility model, stratified by age classes.
The points corresponds to the mean estimate and the vertical bars to the 95%CrI.
(B) Baseline conditional number of work-related contacts as estimated by various prepandemic data and by the mobility model, stratified by age classes.

\newpage

# Data sharing statement

# Acknowledgements

\newpage

# References


