---
title: "COVID-19 disrupted patterns of cause-specific mortality in Switzerland"
date: "`r format(Sys.time(), 'Report generated on Date: %Y-%m-%d Time: %H:%M')`"
author: "Anthony Hauser* (1), Moritz Wagner (2), Anne Laube (2), Julien Riou (1)"
output:
    word_document: 
      reference_docx: template.docx
      number_sections: true
      df_print: paged
bibliography: bibliography.bib
csl: ieee.csl
---

(1) Department of Epidemiology and Health Systems, Unisanté, Center for Primary Care and Public Health and University of Lausanne, Lausanne, Switzerland;<w:br/>
(2) Federal Office of Public Health, Bern, Switzerland;<w:br/>
(*) Corresponding author: anthony.hauser@unisante.ch

```{r setup, include=TRUE}
#Setup
Sys.setlocale("LC_TIME", "C")
wd = getwd()
code_root_path = paste0(strsplit(wd, split="/manuscript")[[1]][1],"/")
print(code_root_path)
source(paste0(code_root_path,"R/000_setup.R"))
knitr::opts_chunk$set(echo = FALSE, warning =FALSE, message=FALSE)

wd_ofsp = "L:/UNISANTE_DESS/S_SUMAC/OFSP_2023/"
data_folder = paste0(wd_ofsp,"02_data/cause_of_death/")
save.date="20241218"
```


```{r, include= TRUE}
#Individual dataset
cod_ind_df = readRDS(paste0(code_root_path,"savepoint/cod_ind_df.RDS"))

causes2 = c("Cardiovascular Diseases","Infectious and Parasitic Diseases",
           "Respiratory Diseases", "Mental and Neurological Disorders",
           "COVID-19",
           "Neoplasms (Cancers)","Suicide","External Causes")

causes2_df = data.frame(cod_group=c("Cardiovascular Diseases","Infectious and Parasitic Diseases",
                                    "Respiratory Diseases", "Mental and Neurological Disorders",
                                    "COVID-19",
                                    "Neoplasms (Cancers)","Suicide","External Causes","Other Causes"),
                        cod_group_label=c("Cardiovascular","Infectious/Parasitic",
                                    "Respiratory", "Mental/Neurological",
                                    "COVID-19",
                                    "Cancers","Suicide","External","Other"),
                        order=c(1,7,4,3,9,2,6,5,8),
                        example = c("Ischaemic heart disease, heart attack", "Sepsis",
                                   "Pulmonary disease, pneumonia, influenza", "Dementia, Alzheimer, Parkinson",
                                   "COVID-19",
                                   "Lung, breast, prostate neoplasms","Intentional self-harm/poisoning","Fall, vehicule accident",
                                   "Unspecified, diabetes, senility, organ disease")) %>%
  #add chapter of cod-10
  left_join(cod_ind_df %>%
              filter(outcome=="ENDG_U_CD_GES_T") %>%
              dplyr::select(cod_group,chapter) %>% unique() %>%
              arrange(cod_group,chapter) %>%
              group_by(cod_group) %>%
              dplyr::summarise(chapter=paste(chapter, collapse=", ")),by="cod_group") %>%
  arrange(order)
```

```{r, include=FALSE}
#Aggregated dataset and results
add_apo = function(x){format(x, big.mark=",")}

cod_agg_pop_df = readRDS(paste0(code_root_path,"savepoint/cod_agg_pop_df.RDS"))
age_classes = cod_agg_pop_df$age_class %>% unique()
setwd(code_root_path)
res_list = load_results_mod6(age_classes, save.date=save.date,mod="mod8")

cod_agg_pop_df = cod_agg_pop_df %>%
  dplyr::mutate(date=ISOweek2date(paste0(cal_year,"-W",ifelse(cal_week<10,paste0("0",cal_week),cal_week),"-1"))) %>% 
  mutate(covid_phase = map2_dbl(date, list(covid_phase), function(d, phases) {
    phase <- phases %>%
      filter(d >= start_date & d <= end_date) %>%
      pull(phase)
    if (length(phase) == 0) NA_real_ else phase
  }))
```


# Abstract

```{r, include=FALSE}

```

**Introduction**
  
  
**Methods and Findings**
  
  
**Conclusion**
  
  
\newpage

# Introduction

Link between covid and other causes
direct: 
indirect: measures -> suicides, respiratory diseases, accident

-> correlation of mortality causes
  -> important to analyse them simultaneously (as increase of a cause can lead to decrease of another cause, and thus no all-cause excess)
  
Disruption of mortality patterns offers a unique opportunity to explore the nature/origin of correlation between causes (which almost all usually peaks during winter but why? is it due to the conditions? to resp virus? other?)

Gap: cause-specific mortality during covid


Objectives

\newpage

# Methods

## Data

Mortality data:
which causes we take
how they were collected
how we aggregated them (definition of groups)
unit: geographical, age, sex, week

Population data

## Model

Infer expected mortality
Model: for each age, time trend, seasonal and overall, effect of sex and region
population size as an offset, log-linear model with poisson distribution
multivariate random effect account for correlation between cause-specific excess mortality rates
between 2011 and 2019

## Excess mortality

Define excess mortality
By cause
All-cause: aggregation of cause-specific excess
by period

## Further analysis

Correlation
Correlation during COVID-19: investigate the partial correlation between cause-specific excess mortality rates during covid-19 between different causes and respiratory disease, which 


Q: where to describe covid-19 phase in switzerland?
\newpage

# Results

## Before 2020

```{r, include=FALSE}

```



Cancer : 
Reclassification of deaths among cancer patients with COVID-19: Many cancer patients who contracted COVID-19 during the 2nd and 3rd waves likely died with contributions from both diseases, but their deaths were primarily attributed to COVID-19 (as per guidelines), reducing recorded cancer mortality. This effect was stronger in 65–79, who were more likely to be hospitalized and tested.
+ harvesting effect of cancer patients occured in wave 2 lead to decrease in deaths in wave 3 
+ absences of resp infections during wave 3

## Excess all-cause
  
## CV, cancer, dementia and infectious or respiratory diseases
  
```{r, include=FALSE}

```



Reasons: 
1) Undiagnosed COVID-19 case : -> increase during wave
2) Long-term mortality displacement -> decrease after waves
3) Absence of resp -> decrease in Jan/Feb
4) Direct consequence of COVID-19: -> increase during wave
5) Short-term mortality displacement -> decrease during wave

Different to say the same
1) Wrong ascertainment : considered as condition (low proportion of condition, among covid deaths) -> increase
5) Wrong ascertainment : considered as covid (high proportion of condition, e.g., cancer, among covid deaths) -> decrease

during wave: + 1), + 4), - 5)
after wave: - 2), + 4) (as a long-term consequence of covid, but not really observed in the results)
Winter: - 3)


among 65-79: no correlation during covid between resp (or covid) and mental (as we observe before pandemic), likely because ascertainment of covid deaths changed over time  (e.g., through testing strategies), as shown by time-varying proportion of covid deaths with mental diseases$



TO DO:
look at overall mortality by cause accounting for principal cause, direct cause and underlying condition.
Compare between covid and pre-covid, especially for cancer in order to see whether overall cancer mortality really decreased or whether it's only because of different attribution of cause of deaths.
could inspire from : https://elifesciences.org/reviewed-preprints/93758?utm_source=chatgpt.com

\newpage

# Discussion

**Summary**
  
**Comparison with other studies**

**Natural experiment: synchronicity between flu and CV/dementia**

**Potential explanations for the observed patterns: 3 types of disruption**

**Strengths, limitations, perspectives**

**Conclusion**
  
  
  
  \newpage

# Figures

```{r, fig.width=10,fig.height=10}
fig1 = res_list$Sigma_mat %>% 
  dplyr::mutate(cod_group = factor(cod_group,levels=causes2_df$cod_group,
                                   labels=causes2_df$cod_group_label),
                cod_group2 = factor(cod_group2,levels=causes2_df$cod_group,
                                    labels=causes2_df$cod_group_label),
                cod_group_id=as.numeric(cod_group),
                cod_group_id2=as.numeric(cod_group2)) %>% 
  rowwise() %>% 
  dplyr::mutate(est_cri = paste0(scales::percent(est, accuracy = 1),"\n",
                                 "[",scales::percent(lwb, accuracy = 1),",",
                                 scales::percent(upb, accuracy = 1),"]")) %>% 
  filter(cod_group_id<cod_group_id2,age_class=="80+") %>% 
  ggplot(aes(x = cod_group, y = fct_rev(cod_group2), fill = abs(est))) +
  geom_tile() +
  geom_text(aes(label = est_cri),
            color = "black", size = 3) +
  scale_fill_gradient(low = "lightyellow", high = "orangered1",limits=c(0,1),
                      name="Correlation",
                      labels=scales::label_percent(accuracy = 1)) +
  labs(x = "", y = "", fill = "Count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0,size=11),
        axis.text.y = element_text(size=11))+
  scale_x_discrete(position = "top")
fig1
```

```{r, include=FALSE}
pdf(file="fig1.pdf",width=10,height=10)
print(fig1)
dev.off()
```

**Figure 1.** Social contact data and mobility data.
(A) Number of participants by survey wave and by country. The opaque part corresponds to the newly enrolled participations (i.e., contributing for the first time), while the more transparent part corresponds to participants that have already taken part in the survey before.
(B) Age distribution. The colored bars display the number of participants by age classes for the four countries. The grey bars in the backgrounds reflect the age distribution in the general population.
(C) Distribution of the number of survey round participated.
(D) Google work-related mobility data. The mobility levels during the weekdays (Monday-Friday) and weekends (Saturday-Sunday) are averaged weekly.


```{r, fig.width=8,fig.height=6}
fig2=fig1
```

```{r, include=FALSE}
pdf(file="fig2.pdf",width=8,height=6)
print(fig2)
dev.off()
```

**Figure 2.** Survey fatigue and effect of mobility on contacts.
(A) Survey fatigue on the probability of reporting no contact $\theta$. The odds ratio (OR) of reporting zero contact is displayed according to the survey round and country
(B) Survey fatigue on the contact rate $\lambda$. The relative number of contacts (relatively to the contact rate $\lambda$) is displayed according to the survey round and country.
(C) Effect of mobility on the probability of reporting no contact and on the contact rate. The estimated effects are displayed for the four countries and with or without adjusting for long-term trend through Gaussian process (GP).

```{r, fig.width=9,fig.height=12}
fig3=fig1
```

```{r, include=FALSE}
pdf(file="fig3.pdf",width=9,height=12)
print(fig3)
dev.off()
```

**Figure 3.** Model fits for the UK.
(A) Probability of having contacts at work, as estimated by the survey model (red points with vertical lines) and as predicted by the mobility model (colored lines). We shows the predictions of three versions of the mobility model: 1) considering the full period (black), 2) adjusting for the long-term change of contact rate (blue) and 3) limiting to the period until February 2021 (green). For the probability of having contacts at work, we do not show the results of the full period without adjustment of the long-term change, as it is almost indistinguishable from the analysis when adjusting for it. The red points corresponds to the mean posterior estimates by survey wave, and the vertical line shows the related 95% credibility interval (95%CrI).
The colored lines correspond to the (weekly) mean posterior estimates and the gray area corresponds to the 95%CrI. The exact definition of the posterior estimates is given in Supplementary S1.3.2.
(B) Contact rate at work as estimated by the survey model (red) and as predicted by the mobility model (black).
(C) Probability of having contacts at work as estimated by the survey model (red) and as predicted by the mobility model (black), according to mobility level. The different levels of darkness in the red estimates of the survey model depicts the different survey waves.


```{r, fig.width=8,fig.height=6}
fig4 = fig1
```

```{r, include=FALSE}
pdf(file="fig4.pdf",width=8,height=6)
print(fig4)
dev.off()
```

**Figure 4.** Baseline estimates.
(A) Baseline proportion of individuals having work-related contacts as estimated by various prepandemic data and by the mobility model, stratified by age classes.
The points corresponds to the mean estimate and the vertical bars to the 95%CrI.
(B) Baseline conditional number of work-related contacts as estimated by various prepandemic data and by the mobility model, stratified by age classes.

\newpage

# Data sharing statement

# Acknowledgements

\newpage

# References


