---
title: "COVID-19 disrupted patterns of cause-specific mortality in Switzerland"
# date: "`r format(Sys.time(), 'Report generated on Date: %Y-%m-%d Time: %H:%M')`"
author: "Anthony Hauser* (1), Moritz Wagner (2), Anna Fesser† (2), Karim Abawi (3), Anne Laube (2), Garyfallos Konstantinoudis (4), Julien Riou (1)" # "Anthony Hauser* (1),  Moritz Wagner (2), Carla Grolimund (2), Katharina Fehst (3), Karim Abawi (3), Anna Fesser (XX), Katrin Schneider (2), Anne Laube (2), Julien Riou (1)"
output:
    word_document: 
      reference_docx: template.docx
      number_sections: true
      df_print: paged
bibliography: bib1.bib
csl: ieee.csl
---

(1) Department of Epidemiology and Health Systems, Unisanté, Center for Primary Care and Public Health and University of Lausanne, Lausanne, Switzerland;<br/>
(2) Federal Statistical Office, Neuchâtel, Switzerland;<br/>
(3) Federal Office of Public Health, Communicable Diseases Division, Bern, Switzerland;<br/>
(4)	Grantham Institute for Climate Change and the Environment, Imperial College London, London, UK;<br/>

(*) Corresponding author: anthony.hauser@unisante.ch<br/>
(†) former affiliation

```{r setup, include=FALSE}
#Setup
Sys.setlocale("LC_TIME", "C")
wd = getwd()
code_root_path = paste0(strsplit(wd, split="/manuscript")[[1]][1],"/")
print(code_root_path)
source(paste0(code_root_path,"R/000_setup.R"))
knitr::opts_chunk$set(echo = FALSE, warning =FALSE, message=FALSE)

wd_ofsp = "L:/UNISANTE_DESS/S_SUMAC/OFSP_2023/"
data_folder = paste0(wd_ofsp,"02_data/cause_of_death/")
save.date="20241218"
```


```{r, include= FALSE}
#Individual dataset
# cod_ind_df = readRDS(paste0(code_root_path,"savepoint/cod_ind_df.RDS"))
# 
# causes2 = c("Cardiovascular Diseases","Infectious and Parasitic Diseases",
#            "Respiratory Diseases", "Mental and Neurological Disorders",
#            "COVID-19",
#            "Neoplasms (Cancers)","Suicide","External Causes")
# 
# causes2_df = data.frame(cod_group=c("Cardiovascular Diseases","Infectious and Parasitic Diseases",
#                                     "Respiratory Diseases", "Mental and Neurological Disorders",
#                                     "COVID-19",
#                                     "Neoplasms (Cancers)","Suicide","External Causes","Other Causes"),
#                         cod_group_label=c("Cardiovascular","Infectious/Parasitic",
#                                     "Respiratory", "Mental/Neurological",
#                                     "COVID-19",
#                                     "Cancers","Suicide","External","Other"),
#                         order=c(1,7,4,3,9,2,6,5,8),
#                         example = c("Ischaemic heart disease, heart attack", "Sepsis",
#                                    "Pulmonary disease, pneumonia, influenza", "Dementia, Alzheimer, Parkinson",
#                                    "COVID-19",
#                                    "Lung, breast, prostate neoplasms","Intentional self-harm/poisoning","Fall, vehicule accident",
#                                    "Unspecified, diabetes, senility, organ disease")) %>%
#   #add chapter of cod-10
#   left_join(cod_ind_df %>%
#               filter(outcome=="ENDG_U_CD_GES_T") %>%
#               dplyr::select(cod_group,chapter) %>% unique() %>%
#               arrange(cod_group,chapter) %>%
#               group_by(cod_group) %>%
#               dplyr::summarise(chapter=paste(chapter, collapse=", ")),by="cod_group") %>%
#   arrange(order)
```

```{r, include=FALSE}
#Aggregated dataset and results
add_apo = function(x){format(x, big.mark=",")}

cod_agg_pop_df = readRDS(paste0(code_root_path,"savepoint/cod_agg_pop_df.RDS"))
age_classes = cod_agg_pop_df$age_class %>% unique()
setwd(code_root_path)
res_list = load_results_mod6(age_classes, save.date=save.date,mod="mod8")

cod_agg_pop_df = cod_agg_pop_df %>%
  dplyr::mutate(date=ISOweek2date(paste0(cal_year,"-W",ifelse(cal_week<10,paste0("0",cal_week),cal_week),"-1"))) %>% 
  mutate(covid_phase = map2_dbl(date, list(covid_phase), function(d, phases) {
    phase <- phases %>%
      filter(d >= start_date & d <= end_date) %>%
      pull(phase)
    if (length(phase) == 0) NA_real_ else phase
  }))
```


# Abstract

```{r, include=FALSE}

```


<!-- The COVID-19 pandemic has disrupted mortality patterns far beyond deaths directly attributed to SARS-CoV-2 infection. -->
<!-- To understand these disruptions, we analysed Swiss weekly deaths (2011-2021) aggregated by age, sex, region and nine groups of causes.  -->
<!-- For each age group, we fitted a Bayesian multivariate Poisson model to 2011–2019 data, accounting for seasonality, long-term trends, and a multivariate random effect capturing cross-cause dependence. -->
<!-- We then compared these predictions with observed 2020–2021 counts to derive age- and cause-specific death excesses. -->
<!-- Cardiovascular deaths spiked during the autumn 2020 wave, whereas none of the studied causes showed sustained excess over 2020-2021. -->
<!-- In those aged ≥80 years, non-COVID respiratory (–25 %) and mental/neurological (–12 %) deaths decreased, largely offset by recorded COVID-19 deaths, leaving a modest +4.5 % all-cause excess. -->
<!-- Cardiovascular and mental/neurological excesses correlated positively with excess mortality attributed to COVID-19 and other respiratory diseases, echoing strong pre-pandemic cross-cause correlations ($\rho>0.80$). -->
<!-- These findings suggest that mortality disruptions during the pandemic likely reflect three overlapping mechanisms: reduced circulation of non-COVID-19 respiratory pathogens, unrecognised COVID-19 deaths (especially cardiovascular), and mortality displacement; and point to a possible underestimation of the true respiratory burden. -->

The COVID-19 pandemic disrupted mortality patterns beyond deaths directly attributed to SARS-CoV-2.
To understand these disruptions, we analysed Swiss weekly deaths (2011–2021) by age, sex, region, and nine cause groups.
For each age group, a Bayesian multivariate Poisson model was fitted to 2011–2019 data, accounting for seasonality, long-term trends, and cross-cause dependence.
Comparing predictions with 2020–2021 observations revealed age- and cause-specific excesses.
Cardiovascular deaths peaked during the autumn 2020 wave, while no cause showed sustained excess across 2020–2021.
In individuals aged ≥80 years, non-COVID respiratory (–25%) and mental/neurological (–12%) deaths declined, largely offset by COVID-19 deaths, leaving a modest +4.5% all-cause excess.
Cardiovascular and mental/neurological excesses correlated with COVID-19 and other respiratory excesses, reflecting strong pre-pandemic cross-cause correlations ($\rho>0.80$).
These patterns likely reflect three overlapping mechanisms: reduced circulation of non-COVID-19 respiratory pathogens, unrecognized COVID-19 deaths (especially cardiovascular), and mortality displacement, pointing to possible underestimation of the true respiratory burden.



<!-- The COVID-19 pandemic has disrupted mortality in many aspects, its effects extending well beyond the direct mortality occuring after SARS-CoV-2 infection. -->
<!-- We aimed to quantify these mortality disruptions for several cause of deaths and understand the different mechanisms behind them. -->
<!-- We used cause-specific data aggregated by week, age, sex, region and nine ICD-10 cause-of-deaths groups from 2011 to 2021. -->
<!-- For each age group, we fitted a Bayesian multivariate Poisson model to 2011–2019 data, including seasonality, long-term trends, and a multivariate random effect capturing cross-cause dependence. -->
<!-- We then compared these predictions with observed 2020–2021 counts to derive age- and cause-specific excesses. -->
<!-- The largest intermittent excess occurred during the autumn 2020 wave: cardiovascular deaths rose by 14\% and 11\% in ages 65–79 and ≥80, respectively. -->
<!-- While cardiovascular deaths show excess duirng the autumn 2020 wave, no cause show sustained excess over 2020-21. -->
<!-- Respiratory (–25\%) and mental/neurological (–12\%) deaths in those ≥80 even showed net deficits. -->
<!-- These were largely offset by COVID-19 deaths, resulting in a net all-cause excess of +4.5\% in the ≥80 population. -->
<!-- Positive partial correlations appeared in 2020–21 between cardiovascular or mental/neurological excess mortality and both COVID-19 and non-COVID respiratory excess mortality. -->
<!-- Similarly, the model inferred strong correlations ($\rho > 0.80$) between respiratory, cardiovascular, and mental/neurological excess deaths during the pre-pandemic period (2011–2019). -->
<!-- These patterns point to three concurrent mechanisms—reduced circulation of other respiratory pathogens, misclassified COVID-19 deaths (especially as cardiovascular), and mortality displacement—suggesting that the true respiratory burden is still underestimated. -->

<!-- We also investigated correlations of excess mortality between causes both before and during the COVID-19 pandemics. -->


<!-- **Introduction** -->

<!-- The COVID-19 crisis reshaped mortality through intertwined direct and indirect pathways. Reported COVID-19 deaths capture only part of the burden because of under-ascertainment, while control measures, infection sequelae, mortality displacement, among other factors, may have altered deaths from other causes. Cause-specific data are therefore indispensable. We quantified cause-specific excess mortality in Switzerland in 2020–21 and examined how excess deaths from different causes were related during the pandemic. -->
<!-- Cause-specific mortality data provide essential insights into how the COVID-19 crisis disrupted mortality patterns across age groups and causes. Multiple mechanisms likely contributed to these shifts: (1) reduced transmission of respiratory pathogens due to public health measures, (2) under-ascertainment of COVID-19 deaths, particularly among those attributed to cardiovascular causes, and (3) mortality displacement, especially in deaths from mental and neurological disorders. The strong correlation between respiratory diseases and mortality from other causes—observed both before and during the pandemic—suggests that the full burden of respiratory infections may be underestimated. This warrants further investigation. -->


<!-- **Methods and Findings** -->

<!-- Weekly deaths for 2011–2021 were aggregated by age, sex, region, and nine ICD-10 cause groups. For each age group, we fitted a Bayesian multivariate Poisson model to 2011–2019 data, including seasonality, long-term trends, and a multivariate random effect capturing cross-cause dependence. Excess deaths were generated from the model’s posterior predictive distribution as the difference between observed and expected deaths. For 2020–21, we used partial correlations to assess how weekly excess deaths from non-COVID causes related to COVID-19 and to other respiratory diseases, adjusting for overlap between the two. -->

<!-- The largest intermittent excess occurred during the autumn 2020 Alpha wave: cardiovascular deaths rose by 14\% and 11\% in ages 65–79 and ≥80, respectively. However, no cause showed sustained excess over the full 2020–21 period. Respiratory (–25\%) and mental/neurological (–12\%) deaths in those ≥80 even showed net deficits. These were largely offset by COVID-19 deaths, resulting in a net all-cause excess of +4.5\% in the ≥80 population. Positive partial correlations appeared in 2020–21 between cardiovascular or mental/neurological excess mortality and both COVID-19 and non-COVID respiratory excess mortality. Similarly, the model inferred strong correlations ($\rho > 0.80$) between respiratory, cardiovascular, and mental/neurological excess deaths during the pre-pandemic period (2011–2019). -->

<!-- **Conclusion** -->

<!-- Cause-specific data reveal how multiple mechanisms acted simultaneously: (1) reduced spread of non-COVID respiratory pathogens, (2) under-ascertained COVID-19 deaths, particularly misclassified as cardiovascular deaths, and (3) mortality displacement affecting mental and neurological causes. Persistent cross-cause correlations—before and during the pandemic—indicate that the burden of respiratory infections is likely underestimated. -->


\newpage

# Introduction

The COVID-19 pandemic has had a profound impact on mortality patterns worldwide.
While much attention has focused on deaths directly attributed to SARS-CoV-2 infection, the total mortality burden linked to the pandemic extends well beyond these immediate effects, reflecting multiple interrelated disruptions to mortality patterns occurring over the short, medium and long term.
First, non-pharmaceutical control measures considerably reduced deaths from other respiratory diseases by limiting the circulation of seasonal viruses [@olsen_decreased_2020].
Second, changes in mobility and social behavior also affected deaths from external causes, with fewer accidents due to reduced mobility and a possible influence of psychosocial stress on suicide risk [@lantos_impact_2024].
Third, in some countries, healthcare disruptions delayed diagnoses and treatment, particularly for cancer, potentially increasing mortality from certain chronic illnesses [@maringe_impact_2020].
Fourth, SARS-CoV-2 infection itself may have contributed to deaths attributed to other causes, especially from cardiovascular causes, either through undiagnosed infection at the time of death or long-term complications [@han_excess_2023].
Finally, the COVID-19 pandemic induced mortality displacement, both over time and across causes, by disproportionately affecting frail individuals (older people or patients with poor prognoses) and thereby altering the risk profile of the surviving population [@holleyman_adjusting_2023].
Studies excluding COVID-19-attributed deaths have shown that mortality from other causes sometimes fell below expected levels, highlighting these indirect effects [@riou_direct_2023].

<!-- In this context, understanding how overall mortality evolved—and how it was distributed across causes—is essential for assessing the full toll of the pandemic. -->

<!-- To monitor mortality during COVID-19 pandemic, several metrics have been used. -->
<!-- COVID-19-related mortality provides an immediate but often incomplete picture, as it only considered direct deaths due to COVID-19, which is subject to underascertainment [@riou_direct_2023]. -->

Given the limitations of COVID-19-attributed mortality, which is prone to underascertainment and only captures immediate consequences of infection, other metrics have come to the forefront to measure the burden of infection [@beaney_excess_2020].
All-cause excess mortality, defined as the difference between observed deaths and the counterfactual number that would have been expected in the absence of pandemic - is more robust, as it does not rely on cause-of-death classification and captures the total mortality impact.
However, as a composite measure, it cannot identify whether apparent increases in some causes are offset by decreases in others.
A natural extension is to estimate cause-specific excess mortality, which allows quantifying the contributions of individual causes over time and, when aggregated, could provide more accurate assessment of total excess mortality [@huynh_joint_2024].
Previous studies on cause-specific excess mortality during COVID-19 have documented important shifts, such as increases in cardiovascular deaths alongside declines in respiratory or dementia mortality, with subtantial heterogeneity across countries [@han_excess_2023; @pizzato_relationship_2025; @paglino_cause-specific_2024; @beeks_cause-specific_2024; @nilsson_cause-specific_2024].
Some investigations have also reported temporal or geographic correlations between COVID-19 activity and non-COVID-19 outcomes [@luck_associations_2023; @paglino_cause-specific_2024; @pizzato_relationship_2025].
However, most of these works focused on one or a limited set of causes, making interpretation difficult given that mortality processes are inherently interdependent.
<!-- Add sentence about the fact that some work presents correlation (temporal or geographical), but specific to some causes and limited to covid. -->
Comprehensive assessments that jointly model multiple causes and explicitly account for cross-cause correlations remain scarce. 
<!-- when available these comparisons focused on covid period, ignoring pre-pandemic preventing from understanding disruptions -->

<!-- In particular, correlations between causes arise in two distinct ways: shared temporal synchrony, where causes rise and fall together due to common seasonal or long-term patterns (e.g., higher winter mortality across many causes), and structural cross-cause dependencies, where mortality from one cause directly interacts with another through mechanisms such as misclassification or mortality displacement. -->
<!-- Understanding the impact of the pandemic requires situating cause-specific excess mortality within both of these correlation structures, rather than considering each cause in isolation. -->

Switzerland maintains high-quality mortality statistics through a thorough process of reviewing each death certificate, yielding rich information on underlying, immediate, and concomitant causes suitable for detailed epidemiological analyses [@weitkunat_mortality_2021].
Leveraging these data, we aim to estimate cause-specific excess mortality in Switzerland from 2011 to 2021, with a particular focus on cross-cause correlations.
We implement a Bayesian multivariate model that captures seasonal and long-term trends, accounts for population dynamics, and models correlation across causes through a shared latent random effects structure.
By examining both the pre-pandemic synchrony in cause-specific mortality and its disruption during the pandemic, we aim to identify signals of cause-of-death misclassification, mortality displacement, and indirect effects of COVID-19.
This approach, by highlighting the interdependence of mortality processes, offers new insight into the mechanisms by which the pandemic and the related control measures have reshaped mortality patterns.

<!-- A natural extension is to estimate cause-specific excess mortality, which allows for several advantages. -->
<!-- First, it helps decompose the overall mortality impact into its underlying components, clarifying change in relative contributions over time. -->
<!-- Second, when aggregated, cause-specific estimates can improve the accuracy of total excess mortality estimates [@huynh_joint_2024]. -->
<!-- However, most studies estimating cause-specific excess mortality analyse each cause independently and do not account for correlations across causes [@han_excess_2023; @pizzato_relationship_2025;@paglino_cause-specific_2024;@beeks_cause-specific_2024;@nilsson_cause-specific_2024]. -->
<!-- This limits interpretability, as understanding these dependencies—both before and during the pandemic—is key to identifying the mechanisms driving disrupted mortality patterns. -->

<!-- Link between covid and other causes -->
<!-- direct:  -->
<!-- indirect: measures -> suicides, respiratory diseases, accident -->

<!-- -> correlation of mortality causes -->
<!--   -> important to analyse them simultaneously (as increase of a cause can lead to decrease of another cause, and thus no all-cause excess) -->

<!-- Disruption of mortality patterns offers a unique opportunity to explore the nature/origin of correlation between causes (which almost all usually peaks during winter but why? is it due to the conditions? to resp virus? other?) -->

<!-- Gap: cause-specific mortality during covid -->



<!-- Q: where to describe covid-19 phase in switzerland? Introduction? -->
\newpage

# Results

## Distribution of deaths

In Switzerland, deaths are heavily concentrated among older adults: 88% occurred in individuals aged 65 and over, while fewer than 2% were among people under 40 in 2020-21 (Fig 1A).
Cardiovascular diseases were the leading cause of death overall, followed by cancers and mental or neurological disorders, though the distribution varied by age group (Fig 1B).
Among individuals aged 80 and above, cardiovascular disease accounted for 33% of all deaths.
In contrast, for those aged 40–79, cancer was the predominant cause, responsible for approximately 40% of deaths.
The relative impact of COVID-19 was greatest in older age groups, contributing to 9% of deaths among those aged 65–79 and 12% among those aged 80 and over.

## Prepandemic mortality trends

The model captures the seasonal and long-term trends in cause-specific mortality between 2011 and 2019 (Fig 2A for individuals aged 80+, and Fig S1–S4 for other age groups), while rare episodes of excess mortality are identified, such as in the winters of 2014/15 and 2016/17.
We observe similarities in mortality trends across six different groups of causes, both in the seasonal patterns and in intermittent episodes of excess mortality that often occur synchronously.
For the 80+ age group, Fig 2C highlights this synchrony in timing, showing that the expected mean mortality peak (in black) occurs consistently across causes between January and February.
During the pre-pandemic period (2011–2019), observed peak timings align closely with these expectations.
However, during the pandemic years, we observe a marked shift: mortality peaks occur earlier than expected.
In addition, the model reveals strong positive correlations in cause-specific excess mortality, mostly among individuals aged 80 and over (Fig 2B).
Specifically, excess deaths from respiratory diseases, mental or neurological disorders, and cardiovascular diseases are highly correlated, with pairwise correlations exceeding 80%.

## COVID-19 mortality

The model also captures all-cause mortality between 2011 and 2019, obtained by aggregating the expected cause-specific mortality.
Among individuals aged 65 and over, the impact of the first two COVID-19 waves—-spring 2020 and fall/winter 2020–21—-clearly exceeded any episodes of excess mortality observed between 2011 and 2019 (Fig 3A).
Cumulative excess mortality highlights the substantial increase in deaths during these waves, followed by a gradual mortality deficit in the subsequent weeks (Fig 3B).
When excluding deaths attributed to COVID-19, we also observe that mortality from other causes was lower than expected in 2020–21 for people aged 65 and over.

## Cause-specific excess mortality in 2020-21

Among individuals aged below 64, almost all cumulative mortalities remained within expected levels over 2020–21, suggesting no evidence of excess mortality (Fig 4A\&B and Fig S5-S9).
A slight excess in suicides was observed among 0-17 over 2020-21 (16 [2-28], +54%), but the small number of observed deaths (43) limits interpretation (Fig S5).
In contrast, among individuals aged 65–79, a deficit in deaths due to cancers and mental or neurological disorders was observed in 2020–21, with approximately 5% fewer deaths than expected (Fig 4C).
The largest decrease occurred during the Alpha wave in January–February 2021.
Mortality due to cardiovascular diseases in this group remained within the expected range, although a localized excess was noted during the autumn 2020 wave (+14%).

Similar patterns were seen in the 80+ age group (Fig 4D).
A marked decrease in mortality from mental and neurological disorders was observed in 2020–21 (−12%), with levels below expectation in all phases except the autumn 2020 wave, when a slight increase occurred.
The most pronounced decrease was during the Alpha wave (−24%).
For cancer, the mortality deficit was also concentrated almost entirely during the Alpha wave.
Cumulative mortality due to cardiovascular diseases among the 80+ was within expected levels overall, but showed substantial variation across phases.
Excess mortality was concentrated during COVID-19 waves, while mortality deficits were more common at the beginning of both 2020 and 2021.
Finally, we observed a consistent deficit in deaths due to respiratory diseases (excluding COVID-19) across all age groups, with the magnitude of the deficit increasing with age.
Fig S5-S9 display cause-specific excess mortality estimates by phases for all causes and all age groups.

To better understand the apparent decrease in cancer-related excess mortality in January–February 2021, we examined deaths among individuals aged 65 and over with a cancer diagnosis in 2020–21.
Unlike analyses based solely on the underlying cause of death, this approach also included individuals with cancer who died from other causes.
In contrast to deaths *from* cancer (pink), the total number of deaths with cancer (black) did not decline during January–February 2021 and even increased slightly in the 65–79 age group and more markedly in those aged 80 and over (Fig S10A).
When stratifying deaths with cancer by survival category, we observed that cancers with high survival exhibited the largest increase during this period, particularly among individuals aged 80 and over (Fig S10B).
<!-- We notice that the decrease in deaths *from* cancer was concentrated among cancers with low baseline survival (Fig S10B). -->
<!-- Conversely, among individuals with cancer dying from another cause, the largest excess was observed in those with high-survival cancer type (Fig S10C). -->

## Correlation with respiratory diseases in 2020-21

We examined the temporal correlation between excess mortality from specific causes and excess mortality due to either COVID-19 or other respiratory diseases during 2020–21.
For cardiovascular and mental/neurological diseases, we observed positive partial correlations with COVID-19 mortality, reaching nearly 0.5 among individuals aged 80 and over (Fig 5 and Fig S11).
Similar correlations were also found with other respiratory diseases, reflecting the high correlations inferred by the model for the prepandemic period 2011-2019.
The strongest correlations with COVID-19 were seen at lags of 1 to 3 weeks, meaning that COVID-19-related excess deaths happened slightly after those from cardiovascular or mental/neurological causes.
In contrast, we observed negative correlations between cancer and COVID-19 mortality, particularly among those aged 65–79.

<!-- -> undetected rather than long-term consequence of covid, as peaks occur simultaneously -->

\newpage

# Discussion

## Summary

We modelled cause-specific mortality trends in Switzerland from 2011 to 2019 to assess excess mortality during the COVID-19 pandemic in 2020–2021.
Compared with the prepandemic baseline, we identified several disruptions in mortality patterns.
Cardiovascular deaths rose temporarily during COVID-19 waves, especially among people aged 65 years and older.
By contrast, respiratory-disease mortality declined consistently throughout the pandemic.
Among older adults, deaths attributed to mental/neurological diseases and to cancer fell below expectation in 2021.
These deficits were more than offset by COVID-19 deaths, resulting in overall all-cause excess mortality.
These shifts likely reflect a combination of mechanisms, including mortality displacement, competing risks, and reduced circulation of other pathogens due to public-health measures.
<!-- synchrony and correlation of excess deaths, observed before covid and after -->

## Potential explanations for the observed patterns: 4 types of disruption

Four broad disruptions could explain the observed patterns.
First, under-ascertainment of COVID-19 deaths inflated the counts attributed to other causes.
Modelling work suggests that only 72\% of COVID-19 deaths were identified as such in Switzerland during 2020-21, with even lower proportions in epidemic peaks [@riou_direct_2023].
Because infection with SARS-CoV-2 can trigger cardiovascular complications, through inflammation, thrombosis, and direct cardiac injury, deaths with COVID-19 as an unrecognized concomitant cause likely contributed to the cardiovascular excesses seen in autumn 2020 and autumn 2021.
This is further supported by the high positive correlations between COVID-19 and cardiovascular excesses.
<!-- + correlation with waves:  -->

Second, sequelae of SARS-CoV-2 infection may increase other cause-specific mortality.
Cohort studies have documented a higher individual risk of delayed complications—particularly cardiovascular events—months after infection [@cai_three-year_2024;@xie_long-term_2022].
Population-level analyses show that these delayed effects contributed to sustained excess cardiovascular mortality in 2021 in countries such as the United States [@han_excess_2023] and the United Kingdom [@british_heart_foundation_excess_nodate].
Such effects are difficult to detect in the relatively healthy Swiss population and may overlap with other disruptions—such as misclassified COVID-19 deaths or mortality displacement—complicating interpretations.

Third, mortality displacement, i.e., when deaths occur earlier than they would have in the absence of the COVID-19 pandemic, partly explains the deficit in mortality observed during the first part of 2021.
The clearest signal appears in mental and neurological disorders: among people aged 65 and over, deaths remained below baseline for several weeks after the autumn 2020 and winter 2020-21 COVID-19 waves.
Nordic countries have seen a similar “harvesting” pattern: some individuals with dementia died during COVID-19 waves rather than later, leaving fewer dementia deaths afterward [@nilsson_cause-specific_2024].
While displacement can persist for months in slowly progressing disorders such as dementia, its impact may appear almost immediately when remaining life expectancy is short, as in advanced cancers.
Using multiple cause death data, we found that the observed decline in cancer deaths at the end of 2020 was offset by an increase in cancer patients dying from other causes, mainly COVID-19, suggesting displacement from cancer to COVID-19 due to competing risks.
This interpretation was supported by the analysis stratifying cancers by survival risk, which showed lower increase in mortality among individuals with low-survival cancers, where the risks of dying from cancer or COVID-19 directly compete.
Such results have also been reproduced theoretically through simulation-based modelling work [@hansen_disentangling_2024].
Finally, the observed negative correlation between cancer-related excess mortality and COVID-19 deaths, which was also reported spatially across US states, further supports this interpretation [@luck_associations_2023].

<!-- While displacement can persist for months in disorders that decline slowly, such as dementia, its impact may appear almost immediately when remaining life expectancy is shorter, as in certain advanced cancers. -->
<!-- Using multiple cause (MC) deaths data, we showed that the apparent fall in cancer deaths at the end of 2020 was offset by an increase in cancer patients dying from other causes, mainly COVID-19. -->
<!-- This suggests that competing risk with COVID-19 causes displacement from cancer to COVID-19. -->
<!-- The presence/role of competing risk was confirmed in the analysis grouping cancer types according to survival. -->
<!-- This analysis showed lower MC deaths in individuals with more deadly cancers, where mortality risks of COVID-19 and cancer competes between each other, limiting the increase of mortality in this specific population. -->
<!-- This mechanism has also been reproduced theoretically through simulation-based modelling work [@hansen_disentangling_2024]. -->
<!-- This mechanism also explains the negative temporal correlation between cancer excess mortality and COVID-19 deaths, a pattern likewise observed across US states [@luck_associations_2023]. -->

Finally, pandemic response—both mandated restrictions and voluntary behavior changes— affected mortality in various ways.
Non-pharmaceutical interventions (NPIs) have reduced the circulation of common respiratory pathogens such as influenza, leading to a marked decrease in mortality caused by respiratory diseases in January-February 2021, a period that normally sees peak winter infections.
Because respiratory infections can precipitate deaths from other conditions, their absence likely also reduced mortality from causes that depend on them, especially in adults aged 80 and over.
This mechanism may account for the mental- and neurological-disease-related death-deficits observed in early 2021 among the 65–79 and 80 + age groups.
On the other hand, the implementation of NPIs has been also used to explain the slight decrease in mortality attributed to cancer during the COVID-19 due to delays in diagnosis observed in the UK [@maringe_impact_2020].
However, this effect is unlikely in Switzerland, where oncology care remained accessible and control measures remained comparatively less stringent [@brunner_comparative_2025].
<!-- ascertainment of cases: Bouman -->

## Cross-cause correlation

Our multivariate model reveals two layers of pre-2020 cause-specific mortality correlation: first, most causes share a winter peak across age groups; second, even after removing this seasonal component, excess deaths from respiratory, cardiovascular, and mental-neurological diseases remain tightly coupled.
The correlation is highest in winters with the largest influenza epidemics, implying that a substantial share of the apparent cardiovascular excess is infection-related (either through biologically triggered events or misclassification) [@fares_winter_2013; @warren-gash_influenza_2009].
While extreme weather or pollution could still contribute, previous studies have suggested that they only explain a minor fraction of the observed excess mortality [@andrews_estimating_2025, @nielsen_excess_2011].
Evidence from 2020-21 reinforces this assumption, as excess cardiovascular and mental/neurological deaths correlate with both COVID-19 and non-COVID-19 respiratory excesses during a period when mortality patterns were disrupted.
In this context, the COVID-19 disruption served as a natural experiment, as COVID-19 waves occurred at unusual periods for a respiratory disease and flu waves were almost non-existent in winter 2020-21.
While the exact mechanisms need to be investigated in more depth, this suggests that the contribution of respiratory pathogens on overall mortality has been underestimated.

## Strengths, limitations and perspectives

The study has several strengths.
We studied the COVID-19 disruption, which temporarily decoupled normally synchronized causes of death, to examine how they are linked.
We built a multi-dimensional Bayesian model that captures population change, long-term trends, seasonal patterns, and cross-cause correlation.
Gaussian processes give this model more flexibility than generalized linear models with splines or seasonal ARIMA, resulting in sharper estimates and more realistic uncertainty [@achilleos_excess_2022; @han_excess_2023; @pizzato_relationship_2025;@paglino_cause-specific_2024;@beeks_cause-specific_2024;@nilsson_cause-specific_2024].
Other approaches that have formally modeled cross-cause correlation, used tensors or multi-cause Gaussian processes, but their complexity often prevents a clear understanding of how cause-specific mortalities are related [@zhang_age-related_2022; @huynh_joint_2024].
Our Bayesian framework explicitly propagates uncertainty, a critical feature when excess-mortality estimates depend on counterfactual predictions.
The analysis also benefits from Switzerland’s high-quality death-certificate data, which follow consistent coding guidelines and achieve near-complete coverage.
Weekly aggregation preserves high temporal granularity and allowed us to detect fine-scale synchrony between causes.

<!-- Regarding the data: high quality and granularity: inidivudal -> aggregated by week to keep high granularity, quality ensures by consistence guidelines over time, and full very few missing deaths -->
<!-- multiple analyses: prepandemic, pandemic with correlation allows to dig a bit into how correlation operates -->
<!-- adds another layer of evidence regarding how causes of deaths are intertwined -->
<!-- This was not only made possible by taking advantage of the special period information leverage by thorough model: multivariate (cross-cause random effect) + seasonality + long-term trend + population change -->

The study also has limitations.
First, most analyses rely solely on the unique underlying cause of death, which may overlook mechanisms such as mortality displacement or competing risks.
The additional analysis focusing on cancer illustrates both this limitation and how incorporating multiple cause death data can provide a more complete picture of mortality dynamics.
Models explicitly representing these mechanisms can be combined to achieve a more rigorous and integrated quantification of mortality displacement across time and causes.
Frailty-survival models capture temporal displacement by accounting for unobserved heterogeneity, while competing-risk models capture cross-cause displacement by redistributing deaths among alternatives [@carollo_competing_2024; @holleyman_adjusting_2023].
Misclassification, however, may persist even with richer cause-of-death information.
Second, some model assumptions, such as the static effects across sex, region, and broad age ranges may oversimplify reality.
It also applies to the estimation of excess mortality in the other causes category, which, by definition, comprises a highly heterogeneous set of conditions.
For this reason, we did not present or interpret these results, and the magnitude of the corresponding all-cause excess mortality estimates (Fig. 3) should be interpreted with caution.
Third, reliance on aggregated data raises the risk of ecological bias: correlations seen at the population level, e.g., between COVID-19 and cardiovascular mortality, do not imply causation for individual cases.
Finally, the dataset ends in 2021, limiting our ability to assess the long-term effects of COVID-19 on cause-specific mortality.

In the future, cause-specific mortality data could serve different purposes.
First, it shows how disruption in one cause—such as a respiratory virus epidemic—can cascade into shifts in other mortality patterns, revealing hidden dependencies.
Second, incorporating these dependencies to predict all-cause mortality might improve accuracy.
Third, such data could be used to explain the divergences in excess mortality patterns observed between countries during the COVID-19 pandemic.
More immediately, several extensions are feasible.
We could apply the same framework to other countries to test whether the correlation patterns we observed in Switzerland are generalisable.
Further, stratifying the analysis by socio-economic status would allow us to characterize the mortality burden across different levels of vulnerability.
Finally, extending the series through 2023 and 2024 will clarify whether the different effects responsible for the disruption of mortality persists after 2021.

## Conclusion

The COVID-19 pandemic caused an exceptional disruption in mortality patterns, affecting multiple causes of death simultaneously.
This disruption provided a unique natural experiment to disentangle mechanisms that have historically overlapped, including misclassification, mortality displacement across time and causes, and the effects of control measures.
Our analysis offers a first step toward understanding the links between causes of death that normally evolve in parallel.
The next step will be to integrate multiple-cause-of-death data and explicitly model complex processes such as the harvesting effect and competing risks to achieve a more complete representation of mortality dynamics.

\newpage

# Methods

## Data

We received individual-level data on cause-specific deaths for the years 2011–2021 from the Federal Statistical Office (FSO), which we aggregated by age group (0–17, 18–39, 40–64, 65–79, 80+), sex, NUTS-2 region, calendar week, and calendar year.
The data include information recorded in the death certificate—such as the immediate, underlying, and contributing (concomitant) causes of death, as well as the official underlying cause of death, which is determined by the FSO based on the coding rules of the International Classification of Diseases, 10th Revision (ICD-10) [@world_health_organization_classification_2009].
All causes of death are coded using ICD-10 and were grouped into nine categories for analysis: cardiovascular diseases (Chapter IX), cancers (Chapter II), mental and neurological disorders (Chapters V and VI), respiratory diseases (Chapter X), External causes (Chapter XX), suicide (Chapter XX), infectious and parasitic diseases (Chapter I), other causes (Chapters III, IV, VII, VIII, XI–XIX, XXI, XXII), and COVID-19 (Chapter XXII) (see Table S1).
Yearly population data by sex, NUTS-2 region, and age group for the years 2011–2021 were also provided by the Federal Statistical Office (FSO).
We applied linear interpolation to obtain weekly population estimates.

<!-- In Switzerland, the internal cause-specific mortality database records detailed information on the circumstances of each death and the characteristics of the deceased. -->
<!-- For every death of a Swiss resident, a physician completes a death certificate specifying the immediate, underlying, and contributing causes of death. -->
<!-- The death certificate is then transmitted to the Federal Statistical Office (FSO), where information is anonymized, coded, and entered into the national mortality database -->
<!-- The official underlying cause of death is determined according to the coding rules of the International Classification of Diseases, 10th Revision (ICD-10). -->
<!-- Additional variables — such as age, canton of residence, calendar week, and year of death — are also recorded. -->


## Model

<!-- We modeled the weekly number of deaths by cause for each age group using a Poisson distribution with a multivariate structure.   -->
<!-- The model accounts for changes in population size $N$ over time via an offset term, and includes fixed effects $\beta$ and $\gamma$ for differences in mortality risk by sex and region, respectively. -->
<!-- Temporal variation is captured by two components: a seasonal trend modeled with a periodic Gaussian Process (GP) $f(t)$, and a long-term trend modeled using a GP with an exponentiated quadratic (squared exponential) kernel $g(t)$.   -->
<!-- To account for the correlation between mortality risks from different causes, the model incorporates a multivariate random effect $\epsilon$. -->
<!-- The model reads as follow: -->
<!-- $$y_{i}(t)\sim Poisson(\mu_{i}(t))$$ -->
<!-- with, -->
<!-- $$\log\mu_{irs}(t) = \alpha_i + \beta_{ir} + \gamma_{is} + f(t) + g(t) + \epsilon_i + \log N_{irst}$$ -->
<!-- where $y_{irs}(t) $ is the observed number of deaths due to cause $i$ at week $t$ in region $r$ and sex $s$. -->
<!-- The parameters $\alpha_i$ represents the baseline relative mortality risk. -->

We modeled the weekly number of deaths by cause for each age group using a Poisson distribution with a multivariate structure under a Bayesian framework, considering only the underlying cause of death.
The model accounts for changes in population size $N$ over time via an offset term, and includes fixed effects $\beta$ and $\gamma$ for differences in cause-specific mortality risk by region and sex, respectively.
Temporal variation is captured by two components: a seasonal trend modeled using a periodic Gaussian Process (GP) with an exponentiated sin-squared kernel $f(t)$, and a long-term trend modeled using a GP with an exponentiated quadratic (squared exponential) kernel $g(t)$ [@riutort-mayol_practical_2022], with $t$ representing calendar weeks.
To account for the correlation between mortality risks across causes, the model incorporates a multivariate random effect $\epsilon(t) \sim \mathcal{N}(0,\Sigma)$, where $\Sigma$ is the covariance matrix.
Correlations between causes were modeled with a Cholesky LKJ(1) prior, which imposes no a priori preference on the strength or direction of the correlations.

The model is specified as follows:
$$
y_{irs}(t) \sim \text{Poisson}(\mu_{irs}(t))
$$
with
$$
\log \mu_{irs}(t) = \alpha_i + \beta_{ir} + \gamma_{is} + f_{i}(t) + g_{i}(t) + \epsilon_i(t) + \log N_{irst},
$$
where $y_{irs}(t)$ denotes the observed number of deaths due to cause $i$ in week $t$, region $r$, and sex $s$, and $\alpha_i$ is the baseline log-transformed mortality risk.
Note that even though we use a Poisson likelihood, this model accounts for potential overdispersion through its complex random-effect structure.
<!-- where $ y_{irs}(t) $ denotes the observed number of deaths due to cause $ i $ at week $ t $, in region $ r $ and sex $ s $; $ \alpha_i $ is the baseline log mortality risk for cause $ i $; $ \beta_{ir} $ and $ \gamma_{is} $ are age-specific fixed effects for region and sex, respectively; $ f(t) $ captures the seasonal variation; $ g(t) $ represents the long-term trend; $ \epsilon \sim \mathcal{N}(0,\Sigma)$ is a multivariate random effect capturing correlation across causes (with correlation matrix $\Sigma$); and $ N_{irst} $ is the population at risk, included as a log offset. -->


The model was run independently for each age group and fitted to the weekly number of cause-specific deaths $y_{irs}(t)$ observed between 1 January 2011 and and 31 December 2019.
It was then used to estimate the weekly *expected number of deaths* $\tilde{y}_{irs}(t)$ in 2020 and 2021, defined as the *posterior predictive distribution* after removing the random effect $\epsilon(t)$ from the log mortality risk $\mu_{irs}(t)$.
We removed $\epsilon(t)$ because it is designed to capture *excess mortality*—that is, unexplained or anomalous variation beyond structured trends.
Weekly excess mortality by cause, region and sex $E_{irs}(t)$ was then defined as the difference between observed and expected deaths:
$$E_{irs}(t) := y_{irs}(t) - \tilde{y}_{irs}(t).$$
<!-- where $y_{irs}(t)$ is the observed number of deaths and $\tilde{y}_{irs}(t)$ the predicted number under baseline conditions.   -->
We also reported aggregated estimates of excess mortality (i.e., *all-cause excess mortality*), across regions, across sexes and across time periods.
This is simply obtained by adding posterior samples.

## Further analysis

We examined how excess deaths from specific causes were related over time to excess deaths from COVID-19 during 2020–21.
To do this, we computed partial Pearson correlations between posterior samples of cause-specific excess deaths and COVID-19 deaths, while controlling for excess deaths from other respiratory diseases.
This adjustment aimed to separate the association with COVID-19 from that with other circulating respiratory pathogens.
We also explored the effect of time lags, by shifting the time series of cause-specific excess deaths up to 8 weeks forward and backward, to assess whether the strength of association varied depending on timing.
We performed the same analysis to assess the relationship between cause-specific excess deaths and excess deaths due to (non-COVID) respiratory diseases, this time adjusting for COVID-19 mortality.

To better understand the observed trends in deaths with cancer as the underlying cause, we conducted an additional analysis including all individuals who died *with* cancer, i.e., where cancer was reported either as the underlying (referred to as *deaths from cancer*) or the immediate cause of death.
We computed the number of deaths from cancer to the total number of deaths with cancer over time and examined their distribution by cancer survival rate to assess whether competing risks with COVID-19 might explain some of the observed patterns.
We performed all the analyses in R version 4.4.1 and Stan version 2.35.0 [@carpenter_stan_2017].
All code is available from <https://github.com/anthonyhauser/causes-of-deaths>.


<!-- Cancer :  -->
<!-- Reclassification of deaths among cancer patients with COVID-19: Many cancer patients who contracted COVID-19 during the 2nd and 3rd waves likely died with contributions from both diseases, but their deaths were primarily attributed to COVID-19 (as per guidelines), reducing recorded cancer mortality. This effect was stronger in 65–79, who were more likely to be hospitalized and tested. -->
<!-- + harvesting effect of cancer patients occured in wave 2 lead to decrease in deaths in wave 3  -->
<!-- + absences of resp infections during wave 3 -->

<!-- Reasons:  -->
<!-- 1) Undiagnosed COVID-19 case : -> increase during wave -->
<!-- 2) Long-term mortality displacement -> decrease after waves -->
<!-- 3) Absence of resp -> decrease in Jan/Feb -->
<!-- 4) Direct consequence of COVID-19: -> increase during wave -->
<!-- 5) Short-term mortality displacement -> decrease during wave -->

<!-- Different to say the same -->
<!-- 1) Wrong ascertainment : considered as condition (low proportion of condition, among covid deaths) -> increase -->
<!-- 5) Wrong ascertainment : considered as covid (high proportion of condition, e.g., cancer, among covid deaths) -> decrease -->

<!-- during wave: + 1), + 4), - 5) -->
<!-- after wave: - 2), + 4) (as a long-term consequence of covid, but not really observed in the results) -->
<!-- Winter: - 3) -->


<!-- among 65-79: no correlation during covid between resp (or covid) and mental (as we observe before pandemic), likely because ascertainment of covid deaths changed over time  (e.g., through testing strategies), as shown by time-varying proportion of covid deaths with mental diseases -->




<!-- multivariate structure can still identify it -->


<!-- excess mortality is prediction with high uncertainty: -> we propagate uncertainty -->
<!-- model: able to correct for trend while propagating uncertainty which is key as excess mortality is about prediction and conterfactuals, -->


<!-- - mortality displacement: needs further investigation but difficult with only deaths data -->
<!--   - competing risk not properly handled: mortality displacement -->
<!--   - frailty model : account for heterogeneity inside group -->
<!-- - misclassification (as we use underlying deaths): (note: maybe it's not misclassification but mortality displacement) -->
<!--   - we acknowledge and try to interpret -->
<!--   - we run in depth investigation using other variable to look at this impact -->
<!-- - broad age groups; static regional and sex effect -->
<!-- - ecological level: difficult to interpret correlations (not at all causal) -->
<!-- - analysis stops at 2021: no long-term effect -->
<!-- - only one outcome : -->
<!--   - complete coverage of deaths, while we use underlying deaths, other variables could be used for in-depth investigation (e.g., cancer) -->




<!-- Perspective -->
<!-- - cause of death data analysis is important because: -->
<!--   - understanding mechanisms and depedencies -->
<!--   - predict all-cause mortality, Huynh -->
<!--   - explain differences observed by countries -->
<!-- - what we could do :  -->
<!--   - apply same approach to other countries to validate correlation and possibly assess reporting quality (e.g., regarding resp diseases deaths) -->
<!--   - investigate this by socio-economic status: level of vulnerabilities -->
<!--   - more recent years -->

<!-- cause of death data analysis is important because: -->
<!-- - understanding mechanisms and depedencies -->
<!-- - predict all-cause mortality, Huynh -->
<!-- - explain differences observed by countries -->





<!-- model: able to correct for trend while propagating uncertainty which is key as excess mortality is about prediction and conterfactuals, -->
<!-- accounting for population change and for correlation between cause -->
<!-- data: accurate information, with clear and consistent over time guidelines to code deaths, and with other variables available to dig about conditions (e.g., for cancer, competing risks) -->
<!-- weekly data -> granulate temporal information -->


<!-- main cause of deaths: difficult to assess in some deaths with multiple comorbidities -->
<!-- difficult to interpret numbers in young population (suicide) as deaths concentrate in older -->
<!-- same for effect of regions -->
<!-- aggregation by cause is arbitrary -->
<!-- separate by age/aggregation by age (i.e, ppl within same age group have identical risk) -->

<!-- competing risk models might be useful to account for mortality displacement over causes -->




<!-- 1) Undetected COVID-19 deaths -->
<!-- synchrony observed between cardiovascular excess and covid waves -->
<!-- -> undetected covid, attributed to cvd, makes senses as covid directly alters cardiovascular system -->
<!-- (-> could be also consequence of covid but less likely as occurs same time) -->
<!-- mental/neurological among 80+ -->

<!-- 2) Effect of covid-19 infection (indistinguishable from undetected covid-19 deaths if the effect is short-term) -->
<!-- Cai et al. -->
<!-- post-acute sequelae condition with CVD symptoms higher after infection, mental/neurological incidence rate  higher, short-term -->
<!-- also higher all-cause excess deaths after adjusting for potential confounders -->

<!-- Fares -->
<!-- Warren-Gash -->
<!-- Han -->

<!-- 3) Deficit in mortality in 2021 can be explained by also mortality displacement -->
<!-- long-term: cvd, dementia (harvesting), died from covid instead of cancer -->
<!-- shorter term: cancer (due to competive risks), die -->
<!-- cvd in 2021, dementia -->
<!-- peak in mortality observed in autumn 2020 (either due to covid-19 or other causes) means that some of the deaths would have occurred anyway, harvesting -->

<!-- competing risk in cancer -> more impact among 65-79, where mortality rate between cancer and covid are more similar -->

<!-- 4) Impact of control measures (indirect impact of covid) -->
<!-- 4b) Care disruption: unlikely, as shown by other study for cancer -->
<!-- or underdiagnosis -> leading to undetected cancer deaths -->
<!-- 4a) Specific deficit in Jan-Feb explained by resp absence (short term) -->
<!-- correlated with respiratory, suggesting an effect -->
<!-- Hygiene measure: absence of mortality peak due to respiratory disease. -->



<!-- Correlation -->

<!-- Two levels : 1) Mortality, 2) Excess mortality, exceptional event -->
<!-- first level could be due to confounders, shared drivers (cold, winter -> confinement inside) -->
<!-- second level makes it more difficult to believe confounder, as they stay constant -->

<!-- some mechanisms (undetected covid-19 deaths, short-term effect) also explains the correlation of cause of deaths during covid -->
<!-- positive for cvd and  -->
<!-- negative for cancer: also observed by Luck -->
<!-- same with other respiratory diseases (maybe don't mention it because difficult to explain) -->
<!-- same as in prepandemic, high correlation with resp mortality (cvd and mental/neurological diseases): 2 levels of correlation (seasonality and excess) -->
<!-- synchrony excess deaths -->
<!-- -> natural experiment during covid (link btw resp and cvd/) -->
<!-- mechanisms/pathways: Fares, Warren-Gash -->




<!-- ## Comparison with other studies -->

<!-- 1) Nilsson, excess mortality by cause in scandinavia -->
<!-- Title: Cause-specific excess mortality in Denmark, Finland, Norway, and Sweden during the COVID-19 pandemic 2020–2022: a study using nationwide population data -->
<!-- model: linear regression -->
<!-- results: excess mortality for cvd -->

<!-- 1) C‑MOR Consortium (~12 countries): Beeks, Achilleos -->
<!-- Title: Cause-Specific Excess Mortality During the COVID-19 Pandemic (2020–2021) in 12 Countries of the C-MOR Consortium -->
<!-- model: two methods, sinusoidal curve to model seasonality or 5 years before pandemic as baseline (Beeks or Achilleos) -->
<!-- multilevel (mixed-effects) -->
<!-- results: conflicting results (btw countries) for cancer, increased mortality for cvd -->
<!-- positive association between covid incidence and cvd excess -->
<!-- negative association between stringency of control measures and cvd (in 2021) -->

<!-- 2) Paglino, SOA -->
<!-- Title: Cause-Specific Excess Mortality During the COVID-19 Pandemic -->
<!-- model: SARIMA often underestimates long-term uncertainty -->
<!-- results: strong correlation between covid-19 deaths and excess from cvd -->
<!-- but it's US, so difficult to compare, still interesting to see that long-term effect (scarring effect) exists there in contrast to CH. -->
<!-- comment: GP more suitable with sudden deviations from seasonal norms -->

<!-- 3) Zhang: in Italy -->
<!-- Title: Bayesian Poisson Regression and Tensor Train Decomposition Model for Learning Mortality Pattern Changes during COVID-19 Pandemic -->
<!-- model: tensor train decomposition, captures latent structure and dependencies in the high dimensional resiudals -->
<!-- regional temporal clustering, cause of deaths patterns -->
<!-- comment: mostly used to identify similar temporal or geographical patterns, difficult to identify association between causes -->

<!-- 4) Han: -->
<!-- Title: Excess cardiovascular mortality across multiple COVID-19 waves in the United States from March 2020 to March 2022 -->
<!-- model: log-linear model with linear temporal trend and periodic spline -->
<!-- results: excess mortality for cvd, due to cardiovascular complication after sars-cov-2 infection and long-term elevated cvd risk, and to reduced cvd care services. -->

<!-- 5) Huynh:  -->
<!-- Title: Joint models for cause-of-death mortality in multiple populations -->
<!-- model: joint modelling age-specific rates of various causes of deaths, Multi-Output Gaussian Processes (MOGP) -->
<!-- results: cannot use them, show how all-cause forecasts benefit from modeling cause interdependencies -->
<!-- comments: very theoretical, no results -->

<!-- 8) Pizzato: correlation between covid and cause-specific excess mortality in 16 countries 2020-23. -->
<!-- Title: Relationship between COVID-19 cases and monthly mortality from all causes, cancer, cardiovascular diseases and diabetes in 16 countries, 2020–21 -->
<!-- model: generalized linear regression with linear trend and seasonal trend with splines, spline also for age -->
<!-- results: positive correlation btw cvd and covid in most countries, no excess for cancer (sometimes lower than expected) -->

<!-- 6) Luck: age-standardize deaths -->
<!-- Title: Associations between mortality from COVID-19 and other causes: A state-level analysis -->
<!-- model: correlation across states between cause of deaths with covid -->
<!-- results: negative correlation for cancer -->

<!-- 7) Carollo -->
<!-- Title: Competing risks models with two time scales -->
<!-- comments: only interesting for competing risk not for cause correlation -->


<!-- ## Natural experiment: synchrony between flu and CV/dementia -->

<!-- ## Potential explanations for the observed patterns: 3 types of disruption -->

<!-- ## Strengths, limitations, perspectives -->

<!-- model: able to correct for trend while propagating uncertainty which is key as excess mortality is about prediction and conterfactuals, -->
<!-- accounting for population change and for correlation between cause -->
<!-- data: accurate information, with clear and consistent over time guidelines to code deaths, and with other variables available to dig about conditions (e.g., for cancer, competing risks) -->
<!-- weekly data -> granulate temporal information -->


<!-- main cause of deaths: difficult to assess in some deaths with multiple comorbidities -->
<!-- difficult to interpret numbers in young population (suicide) as deaths concentrate in older -->
<!-- same for effect of regions -->
<!-- aggregation by cause is arbitrary -->
<!-- separate by age/aggregation by age (i.e, ppl within same age group have identical risk) -->

<!-- competing risk models might be useful to account for mortality displacement over causes -->



\newpage

# Figures

```{r, include=FALSE}
load(paste0(code_root_path,"/manuscript/figures_data.RData"))
```

```{r, fig.width=12,fig.height=6}
fig1_1
```

**Figure 1.** Cause-of-death data in 2020-21. A. Number of deaths by age groups in 2020-21. B. Distribution of the deaths occured in 2020-21 over the nine causes of deaths for each age group.

```{r, fig.width=10,fig.height=14}
fig1[[5]]
```

**Figure 2.** Model fit and cross-cause dependencies for the 80+ age group. A. Model fit. Number of weekly deaths observed in 2011-21 (red points) and model estimates (before 2020) and predictions (2020-21) (black line) with 95\% credibility intervals (95\% CrI, gray area), for the six main causes of deaths for the 80+ age group. B. Cross-cause correlation matrix of the excess mortality for 80+ age group. C. Timing of the mortality peak. The color points and squares correspond to observed peak for each winter (we considered the period from July to June of the next year as peak usually occured during winter). The black points and lines represent the posterior estimate of the peak with 95\% CrI.


```{r, fig.width=11,fig.height=12}
fig2
```

**Figure 3.** All-cause mortality. A. Number of weekly all-cause deaths (i.e., aggregated over all causes of deaths) for each age group. The red points correspond to the observed mortality without COVID-19 and the yellow ones with COVID-19. Black lines and shaded areas correspond to the model posterior mean (i.e., expected mortality) with 95\% CrI. B. Cumulative number of excess deaths with (in yellow) or without (in red) COVID-19 deaths, for each age group.


```{r, fig.width=13.5,fig.height=10.6}
fig3
```


**Figure 4.** Cause-specific excess mortality by phase. Black lines and gray area correspond to posterior mean and 95\% CrI of the cumulative (relative and absolute) excess mortality over 2020-21, respectively. The colored points and vertical lines correspond to posterior mean and 95\% CrI of the relative excess mortality by phase, respectively. The phases are delimited  by the dashed vertical lines. Panels A to D represent the different age groups. See Fig S5-S9 for the excess mortality by phase for each cause and each age group.

```{r, fig.width=10,fig.height=10}
fig4
```

**Figure 5.** Partial correlation between excess mortality due to either COVID-19 (red) or to other respiratory diseases (blue, tagged as "Respiratory Diseases") and excess mortality due to cardiovascular diseases, cancers or mental/neurological diseases, for 65-79 and 80+ age groups, considering different lags (in weeks). Solid color lines correspond to the posterior mean and shaded colored area to the 95\% CrI. The vertical dashed lines indicate the lag with the highest absolute partial correlation. A positive lag corresponds to the situation where excess mortality due to respiratory diseases (either COVID-19 or other respiratory diseases) occurs slightly after excess mortality due to the considered cause.

\newpage

# Data sharing statement



# Acknowledgements

\newpage

# References


